---
title: "BCC CosMx spatial analysis and visualization"
Author: "Massimo Andreatta"
date: "2024-06-18"
editor_options: 
  markdown: 
    wrap: 72
---

This script contains function for visualizing CosMx SMI spatial data.
Based on a pre-processed in situ transcriptomics dataset, we visualize multiple categorical and quantitative variables in the tissue, including predicted cell type, meta-program activity, and CAF signature activation.


```{r}
library(Seurat)
library(sf)
library(ggplot2)
library(future)
plan("sequential")
```

#Read in pre-processed data

Pre-process data first, for cell type annotation, with 1-celltype_insitutype.Rmd
```{r}
seu.obj <- readRDS("cache/cosmx_insitutype_anno.rds")

meta.names <- colnames(seu.obj[[]])
meta.names <- gsub(".x$", "", meta.names, perl=T)
colnames(seu.obj[[]]) <- meta.names
```


Include information for FOV distance from the wound
```{r}
seu.obj$Condition2 <- seu.obj$Condition
ind <- which(!is.na(seu.obj$Dist_from_wound))
seu.obj$Condition2[ind] <- paste(seu.obj$Condition[ind], seu.obj$Dist_from_wound[ind], sep="_")
table(seu.obj$Condition2)
```

```{r}
library(RColorBrewer)
palette2 <- seu.obj@misc$palette
```

Helper functions
```{r}
make_title <- function(obj, cells_of_interest) {
  pat <-  unique(obj@meta.data[cells_of_interest,"Patient_ID"])
  cond <- unique(obj@meta.data[cells_of_interest,"Condition"])
  area <- unique(obj@meta.data[cells_of_interest,"Area_ID"])
  ulc <- unique(obj@meta.data[cells_of_interest,"Ulcerated_area"])
  dist <- unique(obj@meta.data[cells_of_interest,"Dist_from_wound"])
  dir <- unique(obj@meta.data[cells_of_interest,"Wound_direction"])
  
  if (is.na(ulc)) {
    title <- sprintf("%s (%i) - %s", pat, use_fov, cond)
  } else {
    title <- sprintf("%s (%i) - %s", pat, use_fov, ulc)
  }
  if (!is.na(dist)) {
    title <- sprintf("%s - %s", title, dist)
  }
  if (!is.na(dir)) {
    title <- sprintf("%s - %s", title, dir)
  }
  return(title)
}

```

Plot cell type for area
```{r}
plot_celltype_region <- function(seu.obj,
                                 patient_id="PatientA",
                                 condition="Wound",
                                 slide=1,
                                 celltype.metadata="celltype",
                                 palette=NA) {
  
  use_slide_metadata <- unique(seu.obj@meta.data$Run_Tissue_name.x)[slide] 
  use_slide_image <- Images(seu.obj)[slide] # Slide desired, as named in images 
  
  
  cells_of_interest <- seu.obj$id[(seu.obj$Patient_ID == patient_id) &
                                    (seu.obj$Condition %in% condition) &
                                    ( seu.obj$Run_Tissue_name.x == use_slide_metadata)] 
  centroid_data <- seu.obj@images[[use_slide_image]]$centroids 
  zoom_fov <- apply(centroid_data@coords[centroid_data@cells %in% cells_of_interest,], 2, range)
  
  xy_ratio = (max(zoom_fov[,1]) - min(zoom_fov[,1]))/(max(zoom_fov[,2]) - 
                  min(zoom_fov[,2]))
  
  # visualize FOV
  p <- ImageDimPlot(seu.obj, fov = use_slide_image, axes = TRUE, size = 0.7,
                     border.color = "black",
                     border.size = 0.01,
                     dark.background = F,
                     group.by = "celltype",
                     cols = palette,
                     coord.fixed = T) +
    theme_void() +
    xlim(zoom_fov[, 2]) +
    ylim(zoom_fov[, 1]) + coord_flip() + theme(aspect.ratio = 1/xy_ratio)
  
  return(p)
}  
```  

Plot MP activity for area
```{r}
plot_mp_region <- function(seu.obj,
                           patient_id="PatientA",
                           condition="Wound",
                           slide=1,
                           col="red",
                           cols=NULL,
                           min.cutoff=0,
                           max.cutoff=0.3,
                           mp="MP7.WoundResp",
                           celltype="Cancer.cells") {
  
  use_slide_metadata <- unique(seu.obj@meta.data$Run_Tissue_name.x)[slide] 
  use_slide_image <- Images(seu.obj)[slide]
  
  cells_of_interest <- seu.obj$id[(seu.obj$Patient_ID == patient_id) & 
                                    (seu.obj$Run_Tissue_name == use_slide_metadata) &
                                    (seu.obj$Condition == condition)] 
  centroid_data <- seu.obj@images[[use_slide_image]]$centroids 
  zoom_fov <- apply(centroid_data@coords[centroid_data@cells %in% cells_of_interest,], 2, range)
  
  Idents(seu.obj) <- "celltype"
  if(is.na(celltype)) {
    cells.plot <- colnames(seu.obj)
  } else {
    cells.plot = WhichCells(seu.obj, idents = celltype)
  }
  
  xy_ratio = (max(zoom_fov[,1]) - min(zoom_fov[,1]))/(max(zoom_fov[,2]) - 
                  min(zoom_fov[,2]))
  
  if (is.null(cols)) {
    cols = c("white",col)
  }
  
  a <- ImageFeaturePlot(seu.obj,
                        fov = use_slide_image,
                        features = mp,
                        cells = cells.plot,
                        border.color = "black",
                        cols = cols,
                        border.size = 0.01,
                        dark.background = F,
                        coord.fixed = T,
                        min.cutoff = min.cutoff, max.cutoff = max.cutoff) +
  xlim(zoom_fov[, 2]) + ylim(zoom_fov[, 1]) + coord_flip() + theme(aspect.ratio = 1/xy_ratio)
  return(a)
  
}

```


Read in cancer MP signatures, and fibroblast signatures
```{r}
mp_file <- "_aux/TableS2_MPs_filtered.csv"

mps <- read.csv(mp_file)
mps <- as.list(mps)

#We can only evaluate genes in the CosMx panel
panel <- rownames(seu.obj)

mps <- lapply(mps, function(x) {
  x[x %in% panel]
})

fb_file <- "_aux/TableS3_wrCAFsignatures.csv"

fb <- read.csv(fb_file)
fb <- as.list(fb)
fb <- lapply(fb, function(x) {
  x[x %in% panel]
})

```

Score meta-programs
```{r}
library(UCell)
seu.obj <- AddModuleScore_UCell(seu.obj, features=c(mps, fb), name="", ncores=8, maxRank = 200)

seu.obj$`CAF_wound-unwound` <- seu.obj$CAF_wound - seu.obj$CAF_unwound.baseline
```

```{r}
# Set your desired image and FOV 

use_fov <- 1 # FOV desired 
use_slide_metadata <- unique(seu.obj@meta.data$Run_Tissue_name.x)[1] # Slide desired, as named in the metadata column ‘Run_Tissue_name’ 
use_slide_image <- Images(seu.obj)[1] # Slide desired, as named in images 


cells_of_interest <- seu.obj$id[(seu.obj$fov == use_fov) & ( seu.obj$Run_Tissue_name.x == use_slide_metadata)] 
centroid_data <- seu.obj@images[[use_slide_image]]$centroids 
zoom_fov <- apply(centroid_data@coords[centroid_data@cells %in% cells_of_interest,], 2, range)

title <- make_title(seu.obj, cells_of_interest)

# visualize FOV
p1 <- ImageDimPlot(seu.obj, fov = use_slide_image, axes = TRUE, size = 0.7,
                   border.color = "black",
                   border.size = 0.1,
                   dark.background = F,
                   group.by = "celltype",
                   cols = palette2,
                   flip_xy = FALSE,
                   coord.fixed = TRUE) +
  theme_void() + theme(aspect.ratio = 1) +
  xlim(zoom_fov[, 2]) +
  ylim(zoom_fov[, 1]) + ggtitle(title)
p1
```

Visualize FOVs for whole patient
```{r fig.height=12, fig.width=8}

p1 <- plot_celltype_region(seu.obj, patient_id = "PatientA",
                           condition = "Ulcerated_nodular", slide=1,
                           palette=palette2)
p1 + ggtitle("PatientA")
ggsave("plots/PatientA_slide1_allFOVs.pdf", plot=p1, height=12, width=10)

p1 <- plot_celltype_region(seu.obj, patient_id = "PatientB",
                           condition = "Ulcerated_nodular", slide=2,
                           palette=palette2)
p1 + ggtitle("PatientB")
ggsave("plots/PatientB_slide2_allFOVs.pdf", plot=p1, height=12, width=10)

```

MP activity for same slide
```{r fig.height=12, fig.width=8}
cols <- c("red","forestgreen","blue","deeppink","darkcyan","orange","darkmagenta")

a <- plot_mp_region(seu.obj, patient_id = "PatientB", condition = "Ulcerated_nodular",
                    slide = 2, mp="MP7.WoundResp", col = cols[7], celltype="Cancer.cells")

a + ggtitle("PatientB")
ggsave("plots/PatientB_slide2_allFOVs_MP7.pdf", plot=a, height=12, width=8)


b <- plot_mp_region(seu.obj, patient_id = "PatientB", condition = "Ulcerated_nodular",
                    slide = 2, mp="MP2.RadMig", col = cols[2],  celltype="Cancer.cells")

b + ggtitle("PatientB")
ggsave("plots/PatientB_slide2_allFOVs_MP2.pdf", plot=b, height=12, width=8)
```

#Loop over all patients and MPs
```{r eval=F}
pats <- unique(seu.obj$Patient_ID)
mp.use <- names(mps)
#fb.use <- names(fb)
fb.use <- NULL

cols <- c("red","forestgreen","blue","deeppink","darkcyan","orange","darkmagenta")
names(cols) <- mp.use

for (p in pats) {
  if (p %in% c("PatientA","PatientC")) {
    slide = 1
  } else {
    slide = 2
  }
  meta <- seu.obj@meta.data[seu.obj$Patient_ID==p,]
  conds <- unique(meta$Condition)
  for (cc in conds) {
    p1 <- plot_celltype_region(seu.obj, patient_id = p,
                           condition = cc, slide=slide,
                           palette=palette2)
    fname <- sprintf("plots/%s_%s_allFOVs.pdf", p, cc)
    ggsave(fname, plot=p1, height=12, width=10)
    
    for (mm in mp.use) {
      p2 <- plot_mp_region(seu.obj, patient_id = p, condition = cc,
                           col = cols[mm],
                    slide = slide, mp=mm, celltype="Cancer.cells")
      
      fname <- sprintf("plots/%s_%s_allFOVs_%s.pdf", p, cc, mm)
      ggsave(fname, plot=p2, height=12, width=10)
    }
    for (mm in fb.use) {
      p2 <- plot_mp_region(seu.obj, patient_id = p, condition = cc,
                    slide = slide, mp=mm, celltype="Fibroblast")
      
      fname <- sprintf("plots/%s_%s_allFOVs_%s.pdf", p, cc, mm)
      ggsave(fname, plot=p2, height=12, width=10)
    }
  }
}

```

And for wound-responding CAF signature
```{r eval=F}
sig.name <- "CAF_wound-unwound"
pats <- unique(seu.obj$Patient_ID)

hist(seu.obj$`CAF_wound-unwound`, breaks=100)

for (p in pats) {
  if (p %in% c("PatientA","PatientC")) {
    slide = 1
  } else {
    slide = 2
  }
  meta <- seu.obj@meta.data[seu.obj$Patient_ID==p,]
  conds <- unique(meta$Condition)
  for (cc in conds) {
    p2 <- plot_mp_region(seu.obj, patient_id = p, condition = cc, col = "darkred",
                         min.cutoff = -0.3, max.cutoff = 0.3,
                         slide = slide, mp=sig.name, celltype="Fibroblast")
    p2 <- p2 + scale_fill_gradient2(low="darkblue",mid="white",high="darkred")
    fname <- sprintf("plots/%s_%s_allFOVs_%s.pdf", p, cc, sig.name)
    ggsave(fname, plot=p2, height=12, width=10)
  }
}
```


Activity for whole area (Patient C)
```{r fig.height=4, fig.width=10}
a <- plot_mp_region(seu.obj, patient_id = "PatientC", condition = "Wound",
                    slide = 1, mp="MP7.WoundResp", celltype="Cancer.cells")
ggsave("plots/PatientC_wound_allFOVs_MP7.pdf", plot = a, height=15, width=12)


b <- plot_mp_region(seu.obj, patient_id = "PatientC", condition = "Wound",
                    slide = 1, mp="CAF_wound", celltype="Fibroblast")
ggsave("plots/PatientC_wound_allFOVs_wCAF.pdf", plot=b, height=15, width=12)

```

Activity for whole area (Patient D)
```{r fig.height=4, fig.width=10}
a <- plot_mp_region(seu.obj, patient_id = "PatientD", condition = "Wound",
                    slide = 2, mp="MP7.WoundResp", celltype="Cancer.cells")
ggsave("plots/PatientD_wound_allFOVs_MP7.pdf", plot = a, height=15, width=12)


b <- plot_mp_region(seu.obj, patient_id = "PatientD", condition = "Wound",
                    slide = 2, mp="CAF_wound", celltype="Fibroblast")
ggsave("plots/PatientD_wound_allFOVs_wCAF.pdf", plot=b, height=15, width=12)

```


# MP activity for individual FOVs
Make faster by plotting full, and then subsetting plot on each FOV
```{r eval=F}
cols <- c("red","forestgreen","blue","deeppink","darkcyan","orange","darkmagenta")

quants <- sapply(names(mps), function(mp) {
  quantile(seu.obj@meta.data[,mp], probs = 0.99)
})
midpoints <- sapply(names(mps), function(mp) {
  quantile(seu.obj@meta.data[,mp], probs = 0.5)
})


dir.create("plots/MPs_allFOVs")

for (use_slide_metadata in unique(seu.obj$Run_Tissue_name.x)) {
  
  pat <- gsub("\\S+(Patient\\d+)", replacement = "\\1", x = use_slide_metadata, perl=T)
  use_slide_image <- grep(pattern = pat, x = Images(seu.obj), value=T)
  
  #make broad plot first
  pll_area <- lapply(seq_along(mps), function(i) {
     ImageFeaturePlot(seu.obj,
                      fov = use_slide_image,
                      features = names(mps)[i],
                      cells = WhichCells(seu.obj, idents = c("Cancer.cells")),
                      border.color = "black",
                      cols = c("white",cols[i]),
                      border.size = 0.1,
                      dark.background = F,
                      min.cutoff = 0, max.cutoff = quants[i])
   })
  names(pll_area) <- names(mps)
  
  #then focus on individual FOVs
  for (use_fov in unique(seu.obj$fov)) {
    
    cells_of_interest <- seu.obj$id[(seu.obj$fov == use_fov) & ( seu.obj$Run_Tissue_name.x == use_slide_metadata)] 
    centroid_data <- seu.obj@images[[use_slide_image]]$centroids 
    zoom_fov <- apply(centroid_data@coords[centroid_data@cells %in% cells_of_interest,], 2, range)
    
    title <- make_title(seu.obj, cells_of_interest)
    
    
    Idents(seu.obj) <- "celltype"
    
    pll <- lapply(seq_along(mps), function(i) {
      pll_area[[i]] + xlim(zoom_fov[, 2]) + ylim(zoom_fov[, 1]) +
        theme(aspect.ratio = 1) + ggtitle(title) + coord_flip()
    })
    p <- wrap_plots(pll, ncol=4)
    
    ggsave(sprintf("plots/MPs_allFOVs/cosmx_allMPs_%s_fov%i.pdf", use_slide_metadata, use_fov), plot = p, height=8, width=20)
  }
}
```
