---
title: "Meta-program activity in CosMx SMI BCC dataset"
Author: "Massimo Andreatta"
date: "2024-06-18"
editor_options: 
  markdown: 
    wrap: 72
---

Analyze MP activity and CAF signature scores in CosMx FOVs from different conditions.

Conditions: - nodular vs. infiltrative FOVs in nodular-ulcerated samples (locally containing both nodular and infiltrative areas)
            - baseline (day 0) vs. wounded (day7 at same site) vs. unwounded (day7 at adjacent site)


```{r}
library(Seurat)
library(sf)
library(ggplot2)
library(future)
library(RColorBrewer)
plan("sequential")
```

#Read in pre-processed data

Pre-process data first, for cell type annotation, with 1-celltype_insitutype.Rmd
```{r}
seu.obj <- readRDS("cache/cosmx_insitutype_anno.rds")
palette2 <- seu.obj@misc$palette

meta.names <- colnames(seu.obj[[]])
meta.names <- gsub(".x$", "", meta.names, perl=T)
colnames(seu.obj[[]]) <- meta.names
```


Include information for FOV distance from the wound
```{r}
seu.obj$Condition2 <- seu.obj$Condition
ind <- which(!is.na(seu.obj$Dist_from_wound))
seu.obj$Condition2[ind] <- paste(seu.obj$Condition[ind], seu.obj$Dist_from_wound[ind], sep="_")
table(seu.obj$Condition2)
```

Read in cancer MP signatures, and fibroblast signatures
```{r}
mp_file <- "_aux/TableS2_MPs_filtered.csv"

mps <- read.csv(mp_file)
mps <- as.list(mps)

#We can only evaluate genes in the CosMx panel
panel <- rownames(seu.obj)

mps <- lapply(mps, function(x) {
  x[x %in% panel]
})

fb_file <- "_aux/TableS3_wrCAFsignatures.csv"

fb <- read.csv(fb_file)
fb <- as.list(fb)
fb <- lapply(fb, function(x) {
  x[x %in% panel]
})

```

Score meta-programs
```{r}
library(UCell)
seu.obj <- AddModuleScore_UCell(seu.obj, features=c(mps, fb), name="", ncores=8, maxRank = 200)
```

Also calculate differential signatures: MP7 minus MP2, and CAF_wound minus CAF_unwound
```{r}
seu.obj$`MP7-MP2` <- seu.obj$MP7.WoundResp - seu.obj$MP2.RadMig
seu.obj$`CAF_wound-unwound` <- seu.obj$CAF_wound - seu.obj$CAF_unwound.baseline
```

# Differences in nodular vs. infiltrative FOVs (ulcerated-nodular samples)
Analyze MP activity in CosMx FOVs from nodular vs. infiltrative FOVs

```{r}
sub <- subset(seu.obj, subset = Condition == "Ulcerated_nodular")
sub <- subset(sub, subset = Morphology %in% c("Nodular","Infiltrative"))

cancer <- subset(sub, subset = celltype == "Cancer.cells")
fibro <- subset(sub, subset = celltype == "Fibroblast")
```

Average by FOV (cancer cells)
```{r}
library(ggpubr)
min.cells <- 200

means <- aggregate(cancer@meta.data[, names(mps)], list(fov = cancer$pat_fov, morph = cancer$Morphology), mean)
npoints <- aggregate(cancer@meta.data[, 1], list(fov = cancer$pat_fov, morph = cancer$Morphology), length)

means <- means[npoints$x > min.cells, ]

melt <- reshape2::melt(means)
colnames(melt) <- c("fov", "Morphology", "Metaprogram", "Average_UCell_score_by_FOV")

p1 <- ggplot(melt, aes(x=Metaprogram, y=Average_UCell_score_by_FOV, fill=Morphology)) + 
  geom_boxplot() + theme_classic2() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1), axis.title.x = element_blank()) +
  ggtitle("Cancer morphology") +
  scale_fill_manual(values=c("#F44400","#FFFF6C"))  +
  stat_compare_means(label="p.format")
p1 

ggsave("plots/MP_score_by_fov_morphology.pdf", height=4, width=5.3)
```


Split by patient and facet by MP
```{r fig.width=11, fig.height=4}
min.cells <- 200

means <- aggregate(cancer@meta.data[, names(mps)], list(fov = cancer$pat_fov,
                                                        morph = cancer$Morphology,
                                                        pat = cancer$Patient_ID),
                   mean)
npoints <- aggregate(cancer@meta.data[, 1], list(fov = cancer$pat_fov, morph = cancer$Morphology), length)

means <- means[npoints$x > min.cells, ]

melt <- reshape2::melt(means)
colnames(melt) <- c("fov", "Morphology", "Patient", "Metaprogram", "Average_UCell_score_by_FOV")

p <- ggplot(melt, aes(x=Patient, y=Average_UCell_score_by_FOV, fill=Morphology)) + 
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(fill = Morphology), size = 0.8, shape = 21, position = position_jitterdodge(jitter.width = 0.2)) +
  theme_classic2() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1), axis.title.x = element_blank()) +
  ylim(c(0,0.17)) +
  ggtitle("Morphology") +
  facet_wrap(~Metaprogram, ncol = 7) +
  scale_fill_manual(values=c("#F44400","#FFFF6C"))  +
  stat_compare_means(label="p.format", size=3)
p

ggsave("plots/MPs_by_nodinf_faceted.pdf", height=4, width = 11)

```


# Differences in wounded vs. unwounded vs. baseline

Analyze MP activity and CAF signature scores in CosMx FOVs from different conditions:
* baseline (day 0)
* wounded (day7 at same site)
* unwounded (day7 at adjacent site)

```{r}
sub <- subset(seu.obj, subset = Condition %in% c("Baseline","Wound","Unwound"))

cancer <- subset(sub, subset = celltype == "Cancer.cells")
fibro <- subset(sub, subset = celltype == "Fibroblast")
```


On cancer cells, calculate average MP activity by FOV
```{r fig.height=4, fig.width=9}
min.cells <- 200

my_comparisons <- list(c("Baseline", "Wound"),c("Unwound","Wound"))

means <- aggregate(cancer@meta.data[, names(mps)], list(fov = cancer$pat_fov, cond = cancer$Condition), mean)
npoints <- aggregate(cancer@meta.data[, 1], list(fov = cancer$pat_fov, cond = cancer$Condition), length)

means <- means[npoints$x > min.cells, ]

melt <- reshape2::melt(means)
colnames(melt) <- c("fov", "Condition", "Metaprogram", "Average_UCell_score_by_FOV")

p3 <- ggplot(melt, aes(x=Condition, y=Average_UCell_score_by_FOV, fill=Condition)) + 
  geom_boxplot(outlier.shape = NA) + theme_classic2() +
  geom_point(aes(fill = Condition), size = 0.8, shape = 21, position = position_jitterdodge(jitter.width = 0.2)) + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1), axis.title.x = element_blank()) +
  ggtitle("MP activity (cancer cells)") +
  scale_fill_manual(values=c("#07f495","#f4ec07","#f44007")) +
  facet_wrap(~Metaprogram, ncol = 7) +
  stat_compare_means(comparisons = my_comparisons, size=3)
  
p3
```

Include the annotation for "close" vs. "far" from wound
```{r fig.width=12, fig.height=5}
min.cells <- 200

my_comparisons <- list(c("Baseline","Wound_Close"),c("Unwound","Wound_Close"))

means <- aggregate(cancer@meta.data[, names(mps)], list(fov = cancer$pat_fov, cond = cancer$Condition2), mean)
npoints <- aggregate(cancer@meta.data[, 1], list(fov = cancer$pat_fov, cond = cancer$Condition2), length)

means <- means[npoints$x > min.cells, ]

melt <- reshape2::melt(means)
colnames(melt) <- c("fov", "Condition2", "Metaprogram", "Average_UCell_score_by_FOV")
melt$Condition2 <- factor(melt$Condition2, levels=c("Baseline","Unwound","Wound_Far","Wound_Close"))

p3.1 <- ggplot(melt, aes(x=Condition2, y=Average_UCell_score_by_FOV, fill=Condition2)) + 
  geom_boxplot(outlier.shape = NA) + theme_classic2() +
  geom_point(aes(fill = Condition2), size = 0.8, shape = 21, position = position_jitterdodge(jitter.width = 0.2)) + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1), axis.title.x = element_blank()) +
  ggtitle("MP activity (cancer cells)") +
  scale_fill_manual(values=c("#07f495","#f4ec07","orange","#f44007")) +
  facet_wrap(~Metaprogram, ncol = 7) +
  stat_compare_means(comparisons = my_comparisons, size=3)
p3.1
ggsave("plots/MP_score_by_fov_wounding_dist.pdf", height=5, width=12)
```


On CAFs, calculate average signature activity by FOV
```{r}
min.cells <- 50

my_comparisons <- list(c("Baseline", "Wound"),c("Unwound","Wound"))

means <- aggregate(fibro@meta.data[, names(fb)], list(fov = fibro$pat_fov, cond = fibro$Condition), mean)
npoints <- aggregate(fibro@meta.data[, 1], list(fov = fibro$pat_fov, cond = fibro$Condition), length)

means <- means[npoints$x > min.cells, ]

melt <- reshape2::melt(means)
colnames(melt) <- c("fov", "Condition", "CAF_signature", "Average_UCell_score_by_FOV")

p4 <- ggplot(melt, aes(x=Condition, y=Average_UCell_score_by_FOV, fill=Condition)) + 
  geom_boxplot() + theme_classic2() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1), axis.title.x = element_blank()) +
  ggtitle("Fibroblasts (wounding)") +
  scale_fill_manual(values=c("#07f495","#f4ec07","#f44007")) +
  facet_wrap(~CAF_signature, ncol = 2) +
  stat_compare_means(comparisons = my_comparisons, size=3)
  
p4
```

Include the annotation for "close" vs. "far" from wound
```{r}
min.cells <- 50

my_comparisons <- list(c("Baseline","Wound_Close"),c("Unwound","Wound_Close"),c("Wound_Far", "Baseline"))

means <- aggregate(fibro@meta.data[, names(fb)], list(fov = fibro$pat_fov, cond = fibro$Condition2), mean)
npoints <- aggregate(fibro@meta.data[, 1], list(fov = fibro$pat_fov, cond = fibro$Condition2), length)

means <- means[npoints$x > min.cells, ]

melt <- reshape2::melt(means)
colnames(melt) <- c("fov", "Condition2", "CAF_signature", "Average_UCell_score_by_FOV")
melt$Condition2 <- factor(melt$Condition2, levels=c("Baseline","Unwound","Wound_Far","Wound_Close"))

p4.1 <- ggplot(melt, aes(x=Condition2, y=Average_UCell_score_by_FOV, fill=Condition2)) + 
  geom_boxplot() + theme_classic2() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1), axis.title.x = element_blank()) +
  ggtitle("Fibroblasts (wounding)") +
  scale_fill_manual(values=c("#07f495","#f4ec07","orange","#f44007")) +
  facet_wrap(~CAF_signature, ncol = 2) +
  stat_compare_means(comparisons = my_comparisons, size=3)
p4.1

ggsave("plots/CAF_score_by_fov_wounding_dist.pdf", height=4.5, width=6.2)
```



MP7 - MP2
```{r}
min.cells <- 200

means <- aggregate(cancer@meta.data[, "MP7-MP2"], list(fov = cancer$pat_fov, cond = cancer$Condition2), mean)
npoints <- aggregate(cancer@meta.data[, 1], list(fov = cancer$pat_fov, cond = cancer$Condition2), length)
colnames(means) <- c("fov","cond","MP7-MP2")
means <- means[npoints$x > min.cells, ]

melt <- reshape2::melt(means)
colnames(melt) <- c("fov", "Condition", "Metaprogram", "Average_UCell_score_by_FOV")
melt$Condition <- factor(melt$Condition, levels=c("Baseline","Unwound","Wound_Far","Wound_Close"))

pA <- ggplot(melt, aes(x=Metaprogram, y=Average_UCell_score_by_FOV, fill=Condition)) + 
  geom_boxplot() +
  theme_classic2() +
  theme(axis.title.x = element_blank()) +
  ggtitle("Cancer cells (wounding)") +
  scale_fill_manual(values=c("#07f495","#f4ec07","orange","#f44007"))
pA
ggsave("plots/MP_score_by_fov_MP7-MP2.pdf", height=4, width=5.3)
```

CAF differential signature score
```{r fig.height=4, fig.width=4}
min.cells <- 50
my_comparisons <- list(c("Baseline","Wound_Close"),c("Unwound","Wound_Close"),c("Wound_Far", "Unwound"))

means <- aggregate(fibro@meta.data[, "CAF_wound-unwound"], list(fov = fibro$pat_fov, cond = fibro$Condition2), mean)
npoints <- aggregate(fibro@meta.data[, 1], list(fov = fibro$pat_fov, cond = fibro$Condition2), length)
colnames(means) <- c("fov","cond","CAF_wound-unwound")

means <- means[npoints$x > min.cells, ]

melt <- reshape2::melt(means)
colnames(melt) <- c("fov", "Condition", "CAF_signature", "Average_UCell_score_by_FOV")
melt$Condition <- factor(melt$Condition, levels=c("Baseline","Unwound","Wound_Far","Wound_Close"))

pB <- ggplot(melt, aes(x=Condition, y=Average_UCell_score_by_FOV, fill=Condition)) + 
  geom_boxplot(outlier.shape = NA) + theme_classic2() +
  geom_point(aes(fill = Condition), size = 1.5, shape = 21, position = position_jitterdodge(jitter.width = 0.2)) + 
  ggtitle("Fibroblasts (wounding)") +
  scale_fill_manual(values=c("#07f495","#f4ec07","orange","#f44007")) +
  stat_compare_means(comparisons = my_comparisons, size=3) +
  theme(axis.text.x = element_text(angle = 30, vjust = 1, hjust=1), axis.title.x = element_blank())
pB

ggsave("plots/CAF_score_by_fov_Wound-Unwound.pdf", height=4, width=4)
```


Individual genes in CAF signatures
CAF score
```{r}
min.cells <- 50

mat <- LayerData(fibro, layer="data")
matsub <- as.matrix(mat[fb$CAF_wound,])

means <- aggregate(t(matsub), list(fov = fibro$pat_fov, cond = fibro$Condition2), mean)
npoints <- aggregate(fibro@meta.data[, 1], list(fov = fibro$pat_fov, cond = fibro$Condition2), length)

means <- means[npoints$x > min.cells, ]

melt <- reshape2::melt(means)
colnames(melt) <- c("fov", "Condition", "Gene", "Avg_expression")
melt$Condition <- factor(melt$Condition, levels=c("Baseline","Unwound","Wound_Far","Wound_Close"))

ggplot(melt, aes(x=Gene, y=Avg_expression, fill=Condition)) + 
  geom_boxplot() + theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1), axis.title.x = element_blank()) +
  ggtitle("Fibroblasts (wounding)") +
  scale_fill_manual(values=c("#07f495","#f4ec07","orange","#f44007"))

ggsave("plots/Fibro_genes_bywounding.pdf", height=4, width=7.5)
```

