---
title: "Analysis of cell type composition in CosMx SMI"
Author: "Massimo Andreatta"
date: "2024-06-18"
editor_options: 
  markdown: 
    wrap: 72
---

This workbook analyses cell type composition in a CosMx SMI dataset of human BCC. 

Conditions: - nodular vs. infiltrative FOVs in nodular-ulcerated samples (locally containing both nodular and infiltrative areas)
            - baseline (day 0) vs. wounded (day7 at same site) vs. unwounded (day7 at adjacent site)


```{r}
library(Seurat)
library(sf)
library(ggplot2)
library(future)
library(ggpubr)
library(RColorBrewer)
plan("sequential")
```

#Read in pre-processed data

Pre-process data first, for cell type annotation, with 1-celltype_insitutype.Rmd
```{r}
seu.obj <- readRDS("cache/cosmx_insitutype_anno.rds")
seu.obj$annotation_broad <- seu.obj$celltype

meta.names <- colnames(seu.obj[[]])
meta.names <- gsub(".x$", "", meta.names, perl=T)
colnames(seu.obj[[]]) <- meta.names
```


Include information for FOV distance from the wound
```{r}
seu.obj$Condition2 <- seu.obj$Condition
ind <- which(!is.na(seu.obj$Dist_from_wound))
seu.obj$Condition2[ind] <- paste(seu.obj$Condition[ind], seu.obj$Dist_from_wound[ind], sep="_")
table(seu.obj$Condition2)
```

Read in cancer MP signatures
```{r}
mp_file <- "_aux/TableS2_MPs_filtered.csv"

mps <- read.csv(mp_file)
mps <- as.list(mps)

#We can only evaluate genes in the CosMx panel
panel <- rownames(seu.obj)

mps <- lapply(mps, function(x) {
  x[x %in% panel]
})
```

Score meta-programs
```{r}
library(UCell)
seu.obj <- AddModuleScore_UCell(seu.obj, features=mps, name="", ncores=8, maxRank = 200)
```


Define color palette for plotting
```{r}
seu.obj$celltype <- droplevels(seu.obj$celltype)
palette2 <- seu.obj@misc$palette[levels(seu.obj$celltype)]
```

Calculate cell type frequencies by FOV, and log-ratio vs. cancer cell frequency
```{r}
tab <- table(seu.obj$pat_fov, seu.obj$celltype)

freq <- apply(tab, 1, function(x) {x/sum(x)})
logfreq <- apply(tab, 1, function(x) {
  l <- log(x+1)
  l/sum(l)
})

ref = "Cancer.cells"
logratio <- apply(tab, 1, function(x) {
  v <- (1+x)/sum(x+1)
  v <- log2(v/v[ref])
  v[names(v) != ref]
})

```

```{r}
library(dplyr)

freq.m <- reshape2::melt(freq)
colnames(freq.m) <- c("Celltype","pat_fov","value")

cols <- c("pat_fov","Patient_ID","Area_ID","Condition","Condition2","Ulcerated_area","Dist_from_wound","Morphology")

freq.m <- merge(x = freq.m , y = seu.obj@meta.data[,cols],
                     by = "pat_fov") |> distinct()

log.m <- reshape2::melt(logfreq)
colnames(log.m) <- c("Celltype","pat_fov","value")

log.m <- merge(x = log.m , y = seu.obj@meta.data[,cols],
                     by = "pat_fov") |> distinct()


lograt.m <- reshape2::melt(logratio)
colnames(lograt.m) <- c("Celltype","pat_fov","value")

lograt.m <- merge(x = lograt.m , y = seu.obj@meta.data[,cols],
                     by = "pat_fov") |> distinct()

```



1. Compositional differences between nodular and infiltrative FOVs

```{r fig.height=6, fig.width=9}
which.celltypes <- c("MoMacDC","Neutrophil","Tcell","Bcell","PlasmaCell","Mast","Pericyte","Fibroblast","Endothelial","Melanocyte")

this <- freq.m[freq.m$Condition == "Ulcerated_nodular" & freq.m$Morphology %in% c("Infiltrative","Nodular"),]

ggplot(this, aes(x=Celltype, y=value, fill=Morphology)) + geom_boxplot(position="dodge2",outlier.shape = NA) +
  geom_point(aes(fill = Morphology), size = 0.8, shape = 21, position = position_jitterdodge(jitter.width = 0.2)) +
  theme_classic2() + ggtitle("Frequencies") +
  theme(axis.text.x = element_text(angle = 30, hjust=1)) +
  scale_fill_manual(values=c("#F44400","#FFFF6C")) +
  stat_compare_means(label="p.format", size=2)
ggsave("plots/freqs_celltypes_Nod_vs_Inf.pdf", height=4, width=7)

this <- lograt.m[lograt.m$Condition == "Ulcerated_nodular" & lograt.m$Morphology %in% c("Infiltrative","Nodular"),]
this <- this[this$Celltype %in% which.celltypes,]
this$Celltype <- factor(this$Celltype, levels=which.celltypes)

#extend ylim
ylim1 <- min(this$value)
ylim2 <- max(this$value)

ggplot(this, aes(x=Celltype, y=value, fill=Morphology)) + geom_boxplot(position="dodge2",outlier.shape = NA) +
  geom_point(aes(fill = Morphology), size = 0.8, shape = 21, position = position_jitterdodge(jitter.width = 0.2)) +
  theme_classic2() + ggtitle("Log-ratio vs. Cancer cells") +
  theme(axis.text.x = element_text(angle = 30, hjust=1)) +
  scale_fill_manual(values=c("#F44400","#FFFF6C")) +
  ylim(ylim1, ylim2+2) +
  stat_compare_means(label="p.format", size=2)

ggsave("plots/logratios_celltypes_Nod_vs_Inf.pdf", height=4, width=7)
```


2. Compositional differences between baseline vs. wounded vs. unwounded samples

```{r fig.height=8, fig.width=12}
my_comparisons <- list(c("Baseline", "Wound"))
which.celltypes <- c("MoMacDC","Neutrophil","Tcell","Bcell","PlasmaCell","Mast","Pericyte","Fibroblast","Endothelial","Melanocyte")

this <- freq.m[freq.m$Condition %in% c("Baseline","Unwound","Wound"),]
#extend ylim
ylim1 <- min(this$value)
ylim2 <- max(this$value)

ggplot(this, aes(x=Condition, y=value, fill=Condition)) + geom_boxplot(position="dodge2", outlier.shape = NA) +
  geom_point(aes(fill = Condition), size = 0.8, shape = 21, position = position_jitterdodge(jitter.width = 0.2)) +
  theme_classic2() + ggtitle("Frequencies") +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  scale_fill_manual(values=c("#07f495","#f4ec07","#f44007")) +
  ylim(ylim1,ylim2+0.1) +
  facet_wrap(~Celltype, ncol = 7) +
  stat_compare_means(comparisons = my_comparisons)

ggsave("plots/freqs_celltypes_wounded.pdf", height=7.5, width=10)


this <- lograt.m[lograt.m$Condition %in% c("Baseline","Unwound","Wound"),]
this <- this[this$Celltype %in% which.celltypes,]
this$Celltype <- factor(this$Celltype, levels=which.celltypes)
#extend ylim
ylim1 <- min(this$value)
ylim2 <- max(this$value)

ggplot(this, aes(x=Condition, y=value, fill=Condition)) + geom_boxplot(position="dodge2", outlier.shape = NA) +
  geom_point(aes(fill = Condition), size = 0.8, shape = 21, position = position_jitterdodge(jitter.width = 0.2)) +
  theme_classic2() + ggtitle("Log-ratios") +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  scale_fill_manual(values=c("#07f495","#f4ec07","#f44007")) +
  ylim(ylim1,ylim2+1) +
  facet_wrap(~Celltype, ncol = 5) +
  stat_compare_means(comparisons = my_comparisons)
ggsave("plots/logratios_celltypes_wounded.pdf", height=7.5, width=8)

```


2b. Compositional differences between baseline vs. unwounded samples vs. wounded samples (close and far from wound)
```{r fig.height=8, fig.width=12}
my_comparisons <- list(c("Baseline", "Wound_Far"), c("Baseline", "Wound_Close"))
which.celltypes <- c("MoMacDC","Neutrophil","Tcell","Bcell","PlasmaCell","Mast","Pericyte","Fibroblast","Endothelial","Melanocyte")

this <- freq.m[freq.m$Condition %in% c("Baseline","Unwound","Wound"),]
this$Condition2 <- factor(this$Condition2, levels=c("Baseline","Unwound","Wound_Far","Wound_Close"))

#extend ylim
ylim1 <- min(this$value)
ylim2 <- max(this$value)

ggplot(this, aes(x=Condition2, y=value, fill=Condition2)) + geom_boxplot(position="dodge2", outlier.shape = NA) +
  geom_point(aes(fill = Condition2), size = 0.8, shape = 21, position = position_jitterdodge(jitter.width = 0.2)) +
  theme_classic2() + ggtitle("Frequencies") +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  scale_fill_manual(values=c("#07f495","#f4ec07","orange","#f44007")) +
  ylim(ylim1,ylim2+0.3) +
  facet_wrap(~Celltype, ncol = 7) +
  stat_compare_means(comparisons = my_comparisons, label="p.format", size=3)

ggsave("plots/freqs_celltypes_wounded_dist.pdf", height=7.5, width=10)
```
Log-ratios
```{r fig.width=14, fig.height=4}
this <- lograt.m[lograt.m$Condition %in% c("Baseline","Unwound","Wound"),]
this <- this[this$Celltype %in% which.celltypes,]
this$Celltype <- factor(this$Celltype, levels=which.celltypes)
this$Condition2 <- factor(this$Condition2, levels=c("Baseline","Unwound","Wound_Far","Wound_Close"))

#extend ylim
ylim1 <- min(this$value)
ylim2 <- max(this$value)

ggplot(this, aes(x=Condition2, y=value, fill=Condition2)) + geom_boxplot(position="dodge2", outlier.shape = NA) +
  geom_point(aes(fill = Condition2), size = 0.8, shape = 21, position = position_jitterdodge(jitter.width = 0.2)) +
  theme_classic2() + ggtitle("Log-ratios") +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  scale_fill_manual(values=c("#07f495","#f4ec07","orange","#f44007")) +
  ylim(ylim1,ylim2+3) +
  facet_wrap(~Celltype, ncol = 13) +
  stat_compare_means(comparisons = my_comparisons, label="p.format", size=3)
ggsave("plots/logratios_celltypes_wounded_dist.pdf", height=4, width=14)
```