---
title: "BCC CosMx spatial analysis with relation to HH pathway activation"
Author: "Massimo Andreatta"
date: "2024-06-18"
editor_options: 
  markdown: 
    wrap: 72
---

This workbook analyses the activation of hedgehog (HH) pathway genes in an in situ spatial transcriptomics dataset of BCC (CosMx SMI)

```{r}
library(Seurat)
library(sf)
library(ggplot2)
library(future)
library(RColorBrewer)
plan("sequential")
```

#Read in pre-processed data

Pre-process data first, for cell type annotation, with 1-celltype_insitutype.Rmd
```{r}
seu.obj <- readRDS("cache/cosmx_insitutype_anno.rds")
palette2 <- seu.obj@misc$palette

meta.names <- colnames(seu.obj[[]])
meta.names <- gsub(".x$", "", meta.names, perl=T)
colnames(seu.obj[[]]) <- meta.names
```

Include information for FOV distance from the wound
```{r}
seu.obj$Condition2 <- seu.obj$Condition
ind <- which(!is.na(seu.obj$Dist_from_wound))
seu.obj$Condition2[ind] <- paste(seu.obj$Condition[ind], seu.obj$Dist_from_wound[ind], sep="_")
table(seu.obj$Condition2)
```

Helper functions
```{r}
make_title <- function(obj, cells_of_interest) {
  pat <-  unique(obj@meta.data[cells_of_interest,"Patient_ID"])
  cond <- unique(obj@meta.data[cells_of_interest,"Condition"])
  area <- unique(obj@meta.data[cells_of_interest,"Area_ID"])
  ulc <- unique(obj@meta.data[cells_of_interest,"Ulcerated_area"])
  dist <- unique(obj@meta.data[cells_of_interest,"Dist_from_wound"])
  dir <- unique(obj@meta.data[cells_of_interest,"Wound_direction"])
  
  if (is.na(ulc)) {
    title <- sprintf("%s (%i) - %s", pat, use_fov, cond)
  } else {
    title <- sprintf("%s (%i) - %s", pat, use_fov, ulc)
  }
  if (!is.na(dist)) {
    title <- sprintf("%s - %s", title, dist)
  }
  if (!is.na(dir)) {
    title <- sprintf("%s - %s", title, dir)
  }
  return(title)
}

```

Plot cell type for area
```{r}
plot_celltype_region <- function(seu.obj,
                                 patient_id="PatientA",
                                 condition="Wound",
                                 slide=1,
                                 celltype.metadata="celltype",
                                 palette=NA) {
  
  use_slide_metadata <- unique(seu.obj@meta.data$Run_Tissue_name.x)[slide] 
  use_slide_image <- Images(seu.obj)[slide] # Slide desired, as named in images 
  
  
  cells_of_interest <- seu.obj$id[(seu.obj$Patient_ID == patient_id) &
                                    (seu.obj$Condition %in% condition) &
                                    ( seu.obj$Run_Tissue_name.x == use_slide_metadata)] 
  centroid_data <- seu.obj@images[[use_slide_image]]$centroids 
  zoom_fov <- apply(centroid_data@coords[centroid_data@cells %in% cells_of_interest,], 2, range)
  
  xy_ratio = (max(zoom_fov[,1]) - min(zoom_fov[,1]))/(max(zoom_fov[,2]) - 
                  min(zoom_fov[,2]))
  
  # visualize FOV
  p <- ImageDimPlot(seu.obj, fov = use_slide_image, axes = TRUE, size = 0.7,
                     border.color = "black",
                     border.size = 0.01,
                     dark.background = F,
                     group.by = "celltype",
                     cols = palette,
                     coord.fixed = T) +
    theme_void() +
    xlim(zoom_fov[, 2]) +
    ylim(zoom_fov[, 1]) + coord_flip() + theme(aspect.ratio = 1/xy_ratio)
  
  return(p)
}  
```  

Plot MP activity for area
```{r}
plot_mp_region <- function(seu.obj,
                           patient_id="PatientA",
                           condition="Wound",
                           slide=1,
                           col="red",
                           cols=NULL,
                           min.cutoff=0,
                           max.cutoff=0.3,
                           mp="MP7.WoundResp",
                           celltype="Cancer.cells") {
  
  use_slide_metadata <- unique(seu.obj@meta.data$Run_Tissue_name.x)[slide] 
  use_slide_image <- Images(seu.obj)[slide]
  
  cells_of_interest <- seu.obj$id[(seu.obj$Patient_ID == patient_id) & 
                                    (seu.obj$Run_Tissue_name == use_slide_metadata) &
                                    (seu.obj$Condition == condition)] 
  centroid_data <- seu.obj@images[[use_slide_image]]$centroids 
  zoom_fov <- apply(centroid_data@coords[centroid_data@cells %in% cells_of_interest,], 2, range)
  
  Idents(seu.obj) <- "celltype"
  if(is.na(celltype)) {
    cells.plot <- colnames(seu.obj)
  } else {
    cells.plot = WhichCells(seu.obj, idents = celltype)
  }
  
  xy_ratio = (max(zoom_fov[,1]) - min(zoom_fov[,1]))/(max(zoom_fov[,2]) - 
                  min(zoom_fov[,2]))
  
  if (is.null(cols)) {
    cols = c("white",col)
  }
  
  a <- ImageFeaturePlot(seu.obj,
                        fov = use_slide_image,
                        features = mp,
                        cells = cells.plot,
                        border.color = "black",
                        cols = cols,
                        border.size = 0.01,
                        dark.background = F,
                        coord.fixed = T,
                        min.cutoff = min.cutoff, max.cutoff = max.cutoff) +
  xlim(zoom_fov[, 2]) + ylim(zoom_fov[, 1]) + coord_flip() + theme(aspect.ratio = 1/xy_ratio)
  return(a)
  
}

```

Read in cancer MP signatures, and HH signatures
```{r}
mp_file <- "_aux/TableS2_MPs_filtered.csv"

mps <- read.csv(mp_file)
mps <- as.list(mps)

#We can only evaluate genes in the CosMx panel
panel <- rownames(seu.obj)

mps <- lapply(mps, function(x) {
  x[x %in% panel]
})

HHs <- read.csv("_aux/HH_signatures.csv")
HHs <- as.list(HHs)
HHs <- lapply(HHs, function(x) x[x!=""])

#only use genes in the panel
HHs <- lapply(HHs, function(x) {
  x[x %in% panel]
})
```

Score meta-programs
```{r}
library(UCell)
seu.obj <- AddModuleScore_UCell(seu.obj, features=c(mps, HHs), name="", ncores=8, maxRank = 200)

seu.obj$`MP7_MP2` <- seu.obj$MP7.WoundResp - seu.obj$MP2.RadMig
```

```{r}
which.HH <- "KEGG_HEDGEHOG_SIGNALING_PATHWAY"

seu.obj$HHscore <- seu.obj@meta.data[,which.HH]

sigs.score.data <- seu.obj@meta.data[,c(grep("MP",names(seu.obj@meta.data),value = T),
                                    "HHscore","Patient_ID","pat_fov","celltype")]
#only use ulcerated-nodular samples
sigs.score.data <- sigs.score.data[sigs.score.data$Patient_ID %in% c("PatientA","PatientB"),]
#only look at cancer cells for HH pathway
sigs.score.data <- sigs.score.data[sigs.score.data$celltype == "Cancer.cells",]

# number of FOVs
length(unique(sigs.score.data$pat_fov))

head(sigs.score.data)
```

Average correlation coefficient by FOV
```{r}
mp.use <- names(mps)
corr <- lapply(names(mps), function(mp) {
  v <- c()
  for (s in unique(sigs.score.data$pat_fov)) {
    sub <- sigs.score.data[sigs.score.data$pat_fov == s,]
    v[s] <- cor(x=sub[,mp], y=sub[,"HHscore"], method="pearson")
  }
  v
}) 
names(corr) <- mp.use
df <- as.data.frame(corr)

#sort by mean MP activity
avgs <- apply(df, 2, mean)
order <- sort(avgs, decreasing = T)

df$FOV <- rownames(df)
melt <- reshape2::melt(df)
colnames(melt) <- c("FOV","MP","Correlation")
melt$MP <- factor(melt$MP, levels=names(order))

cols <- c("red","forestgreen","blue","deeppink","darkcyan","orange","darkmagenta")
names(cols) <- mp.use

a <- ggplot(melt, aes(y = Correlation, x = MP, fill=MP)) +
    scale_fill_manual(values=cols) +
    geom_hline(yintercept=0, linetype='dotted') +
    geom_boxplot(width=0.8, color="black", outlier.size = 2) +
    theme_classic2() +
    theme(axis.text.x = element_text(angle = 30, vjust = 1, hjust=1), axis.title.x = element_blank()) +
  ggtitle("Correlation with HH score by FOV (N=27)")
a

ggsave("plots/MPs_vs_HH_correlation_by_FOV.pdf", height=4.5, width=4.5)
```


MP activity, and HH score for same slide
```{r fig.height=12, fig.width=8}
which.pat <- "PatientB"
slide <- 2
cols <- c("red","forestgreen","blue","deeppink","darkcyan","orange","darkmagenta")

c <- plot_mp_region(seu.obj, patient_id = which.pat, condition = "Ulcerated_nodular",
                    min.cutoff = -0.2, max.cutoff = 0.2,
                    slide = slide, mp="MP7_MP2", celltype="Cancer.cells") +
  scale_fill_gradient2(low=cols[2],mid="white",high=cols[7])
c + ggtitle(which.pat)
ggsave(sprintf("plots/%s_slide%s_allFOVs_MP7-MP2.pdf", which.pat, slide), plot=c, height=12, width=8)

d <- plot_mp_region(seu.obj, patient_id = which.pat, condition = "Ulcerated_nodular",
                    min.cutoff = 0, max.cutoff = 0.3,
                    slide = slide, mp="HH_targets", col = "brown",  celltype="Cancer.cells")
d + ggtitle(which.pat)

ggsave(sprintf("plots/%s_slide%s_allFOVs_HHsignature_target.pdf", which.pat, slide), plot=d, height=12, width=8)
```
