---
title: "Cell-cell interactions in CosMx BCC dataset"
Author: "Massimo Andreatta"
date: "2024-06-18"
editor_options: 
  markdown: 
    wrap: 72
---

Characterize the cell-cell interactions in a CosMx SMI spatial transcriptomics dataset of human BCC. We look at homotypic cell-cell interactions (i.e. between cells of the same type) vs. heterotypic cell-cell interactions (i.e. between cells of different types); in particular for cancer cells, are tumor samples/areas of a certain condition enriched in homotypic/heterotypic interactions? is the heterotypic on homotypic ratio a good representation of cancer complexity? The we attempt to quantify the interactome in terms of enriched interactions between pairs of cell types.

Conditions: - nodular vs. infiltrative FOVs in nodular-ulcerated samples (locally containing both nodular and infiltrative areas)
            - baseline (day 0) vs. wounded (day7 at same site) vs. unwounded (day7 at adjacent site)



```{r}
library(Seurat)
library(sf)
library(ggplot2)
library(future)
library(RColorBrewer)
library(igraph)
plan("sequential")
```


#Define functions to be used in this workbook

1. Calculate vector of neighbors for each cell, adding a new column to metadata
```{r}
radius_nn <- function(meta, k=50, radius=200, types=NULL,
                       meta.labels="celltype") {
  
  require(RANN)
  
  if (is.null(types)) {
     types <- levels(meta[,meta.labels])
  }
  counts <- table(meta[,meta.labels])
  
  #exclude low quality
  types <- types[!types %in% c("Unknown","Low.quality")]
  
  meta <- meta[meta[,meta.labels] %in% types,]
  
  coords <- meta[,c("x","y")]
  labs <- meta[,meta.labels]
  
  # Find neighbors within the current radius range for each anchor cell
  neighbors <- nn2(coords, k=k, searchtype = "radius", radius = radius)
  #drop match to self
  neighbors$nn.idx <- neighbors$nn.idx[,2:ncol(neighbors$nn.idx)]
  
  neighbors_vec <- apply(neighbors$nn.idx, 1, function(x) {
    paste(x[x!=0],collapse=",")
  })
  meta$neighbors_vector <- neighbors_vec
  
  #Add vector of frequencies for neighbor types
  neighbors_encode <- apply(neighbors$nn.idx, 1, function(x) {
    x <- x[x!=0]
    types <- table(labs[x])
    paste(types,collapse=",")
  })
  meta$neighbors_encoding <- neighbors_encode
  
  meta
}
```

2. Generate a title for spatial plots
```{r}
make_title <- function(obj, cells_of_interest) {
  pat <-  unique(obj@meta.data[cells_of_interest,"Patient_ID"])
  cond <- unique(obj@meta.data[cells_of_interest,"Condition"])
  area <- unique(obj@meta.data[cells_of_interest,"Area_ID"])
  dist <- unique(obj@meta.data[cells_of_interest,"Dist_from_wound"])
  dir <- unique(obj@meta.data[cells_of_interest,"Wound_direction"])

  title <- sprintf("%s (%i) - %s", pat, use_fov, cond)

  if (!is.na(dist)) {
    title <- sprintf("%s - %s", title, dist)
  }
  if (!is.na(dir)) {
    title <- sprintf("%s - %s", title, dir)
  }
  return(title)
}
```

3. For each cell, find the closest neighbor of each cell type
```{r}
closest_celltype <- function(meta, k=3, radius=1000, meta.labels="celltype",
                             min.cells=5, symmetric = TRUE,
                          types = NULL) {
  require(RANN)
  
  if (is.null(types)) {
     types <- levels(meta[,meta.labels])
  }
  counts <- table(meta[,meta.labels])
  types <- types[types %in% names(counts)[counts >= min.cells]]
  
  #exclude low quality
  types <- types[!types %in% c("Unknown","Low.quality")]
  
  meta <- meta[meta[,meta.labels] %in% types,]
  
  coords <- meta[,c("x","y")]
  labs <- meta[,meta.labels]
  
  avgs <- data.frame(matrix(ncol = length(types) , nrow = length(types)), row.names = types)
  sds <- data.frame(matrix(ncol = length(types) , nrow = length(types)), row.names = types)
  colnames(avgs) <- types
  colnames(sds) <- types
  
  for (l1 in types) {
    query <- meta[meta[,meta.labels] == l1,]
    query.c <- query[,c("x","y")]
    for (l2 in types) {
      data <- meta[meta[,meta.labels] == l2,]
      data.c <- data[,c("x","y")]
      
      k.use <- min(k, nrow(data.c))
      # Find neighbors within the current radius range for each anchor cell
      neighbors <- nn2(data=data.c, query=query.c, k=k.use, searchtype = "radius", radius = radius)
      
      min.dist <- apply(neighbors$nn.dists, 1, function(r) {
        r <- r[r>0]
        if (length(r) < 1) {
          min <- radius
        } else {
          min <- min(r)
        }
        min
      })
      
      avgs[l1,l2] <- mean(min.dist)
      sds[l1,l2] <- sd(min.dist)
    }
  }
  
  if (symmetric) {
    for (i in 1:nrow(avgs)) {
      for (j in 1:ncol(avgs)) {
        if (avgs[i,j] > avgs[j,i]) {
          avgs[i,j] <- avgs[j,i]
          sds[i,j] <- sds[j,i]
        }
      }
    }
  }
  
  closest <- list()
  closest$avg <- avgs
  closest$sd <- sds
  return(closest)
}  
```

4. Draw interactome of a sample as a graph
```{r}
make.graph.plot <- function(meta, knn.res, min.value=3, width.wgt=0.2, layout=c("kk","fr")) {
  
  g <- graph_from_adjacency_matrix(as.matrix(1000/knn.res$avg), mode="undirected", weighted=T, diag=F)
  gsd <- graph_from_adjacency_matrix(as.matrix(1000/knn.res$sd), mode="undirected", weighted=T, diag=F)
  
  #E(g)$width <- width.wgt*(E(g)$weight)
  E(g)$width <- width.wgt*(E(gsd)$weight)
  
  E(g)$width[E(g)$weight < min.value] <- 0
  g <- g - E(g)[E(g)$width == 0]
  
  V(g)$color <- palette[V(g)$name]
  
  wgt <- table(meta$celltype)
  wgt <- wgt/sum(wgt)
  wgt <- wgt[colnames(knn.res$avg)]
  V(g)$size <- 100*sqrt(wgt)+1
  
  if (layout[1] == "fr") {
     plot(g, layout=layout_with_fr(g, weight = E(g)$weight))
  } else {
     plot(g, layout=layout_with_kk(g, weight = round(1000/E(g)$weight)))
  }
}
```


#Read in pre-processed data
Pre-process data first, for cell type annotation, with 1-celltype_insitutype.Rmd
```{r}
seu.obj <- readRDS("cache/cosmx_insitutype_anno.rds")
seu.obj$annotation_broad <- seu.obj$celltype
palette <- seu.obj@misc$palette

meta.names <- colnames(seu.obj[[]])
meta.names <- gsub(".x$", "", meta.names, perl=T)
colnames(seu.obj[[]]) <- meta.names
```

Include information for FOV distance from the wound
```{r}
seu.obj$Condition2 <- seu.obj$Condition
ind <- which(!is.na(seu.obj$Dist_from_wound))
seu.obj$Condition2[ind] <- paste(seu.obj$Condition[ind], seu.obj$Dist_from_wound[ind], sep="_")
table(seu.obj$Condition2)
```


# Estimate average cell radius
```{r}
areas <- seu.obj$Area
hist(areas)
q <- quantile(areas)

area_avg <- q["50%"]
radius_avg <-  sqrt(area_avg/pi)
```

#Score MP signatures

Read in cancer MP signatures, and fibroblast signatures
```{r}
mp_file <- "_aux/TableS2_MPs_filtered.csv"

mps <- read.csv(mp_file)
mps <- as.list(mps)

#We can only evaluate genes in the CosMx panel
panel <- rownames(seu.obj)

mps <- lapply(mps, function(x) {
  x[x %in% panel]
})

fb_file <- "_aux/TableS3_wrCAFsignatures.csv"

fb <- read.csv(fb_file)
fb <- as.list(fb)
fb <- lapply(fb, function(x) {
  x[x %in% panel]
})

```

Score meta-programs
```{r}
library(UCell)
seu.obj <- AddModuleScore_UCell(seu.obj, features=c(mps, fb), name="", ncores=8, maxRank = 200)
```

Also calculate differential signatures: MP7 minus MP2, and CAF_wound minus CAF_unwound
```{r}
seu.obj$invasiveness <- seu.obj$MP7.WoundResp - seu.obj$MP2.RadMig
seu.obj$invasiveCAF <- seu.obj$CAF_wound - seu.obj$CAF_unwound.baseline

cancer <- subset(seu.obj, subset=celltype=="Cancer.cells")
caf <- subset(seu.obj, subset=celltype=="Fibroblast")
```


Add X and Y coordinates for cell centroids to metadata
```{r}
fovs <- unique(seu.obj$pat_fov)
imgs <- Images(seu.obj)
diet <- DietSeurat(seu.obj)
diet@images <- list()

meta.list <- lapply(fovs, function(f) {
  
  pat <- gsub("\\S+(Patient\\d+)_.*", replacement = "\\1", x = f, perl=T)
  image <- grep(pattern = pat, x = Images(seu.obj), value=T)
  
  this <- subset(diet, subset=pat_fov == f)
  cells_this <- colnames(this)
  
  centroid_data <- seu.obj@images[[image]]$centroids
  centroid_this <- centroid_data@coords[centroid_data@cells %in% cells_this,]
  rownames(centroid_this) <- centroid_data@cells[centroid_data@cells %in% cells_this]
  
  this <- AddMetaData(this, metadata = centroid_this)
  
  this[[]]
})
names(meta.list) <- fovs
```

Calculate vector of neighbors for each cell, adding a new column to metadata
```{r}
#how many radii to look for neighboring cells
distance_max <- 5*radius_avg

#meta <- meta.list$Run6057_Patient1_1
meta <- Reduce(f=rbind, x=meta.list)

meta <- radius_nn(meta, radius = distance_max)
```



# Calculate homotypic vs. heterotypic cell-cell interactions

Homotypic cell-cell interactions (for cancer cells) are defined as those where the majority (>90%) of neighbors of a given cancer cell are also cancer cells (or normal keratinocytes, or melanocytes); otherwise the interaction is labeled as heterotypic.
```{r}
levs <- levels(meta$celltype)

cancer <- meta[meta$celltype == "Cancer.cells",]

#Define which cell types in neighborhood, and at which min frequency, to consider a cell homotypic
idx <- which(levs %in% c("Cancer.cells", "Normal.Kerat", "Melanocyte"))

min.fq <- 0.9

#cancer cells with only cancer cells as neighbors
cancer.homo <- sapply(cancer$neighbors_encoding, function(x) {
  vec <- as.numeric(strsplit(x, split=",")[[1]])
  sum <- sum(vec[idx])
  tot <- sum(vec)
  
  if (sum > 0 & sum/tot >= min.fq) {
    "homotypic"
  } else {
    "heterotypic"
  }
})
cancer$homotypic <- cancer.homo
```

Homo/heterotypic interaction are only defined for cancer cells here.
```{r}
seu.obj$homotypic <- "non-cancer"
seu.obj@meta.data[rownames(cancer),"homotypic"] <- cancer$homotypic

seu.obj$homotypic <- factor(seu.obj$homotypic, levels=c("homotypic","heterotypic","non-cancer"))

seu.obj <- AddMetaData(seu.obj, metadata = meta[,c("x","y","neighbors_vector","neighbors_encoding")])
```

Include coordinates as a dim reduction
```{r}
matrix <- seu.obj@meta.data[,c("x","y")]
dimred <- as.matrix(matrix)
colnames(dimred) <- c("XY_1","XY_2")
#New dim reduction
seu.obj@reductions[["XY"]] <- new("DimReduc",
                                         cell.embeddings = dimred,
                                         assay.used = "Nanostring",
                                         key = "XY_",
                                         global = FALSE)
```



Make a plot of heterotypic vs. homotypic interactions for a given FOV
```{r fig.height=5, fig.width=12}
use_fov <- 38 # FOV desired 
use_slide_metadata <- unique(seu.obj@meta.data$Run_Tissue_name.x)[1] # Slide desired, as named in the metadata column ‘Run_Tissue_name’ 
use_slide_image <- Images(seu.obj)[1] # Slide desired, as named in images 

cells_of_interest <- seu.obj$id[(seu.obj$fov == use_fov) & ( seu.obj$Run_Tissue_name.x == use_slide_metadata)] 
centroid_data <- seu.obj@images[[use_slide_image]]$centroids 
zoom_fov <- apply(centroid_data@coords[centroid_data@cells %in% cells_of_interest,], 2, range)

title <- make_title(seu.obj, cells_of_interest)

# visualize FOV using default settings (no cell boundaries)
p1 <- ImageDimPlot(seu.obj, fov = use_slide_image, axes = TRUE, size = 0.7,
                   border.color = "black",
                   border.size = 0.1,
                   dark.background = F,
                   group.by = "celltype",
                   cols = palette,
                   flip_xy = FALSE,
                   coord.fixed = TRUE) +
  theme_void() + theme(aspect.ratio = 1) +
  xlim(zoom_fov[, 2]) +
  ylim(zoom_fov[, 1]) + ggtitle(title)

p2 <- ImageDimPlot(seu.obj, fov = use_slide_image, axes = TRUE, size = 0.7,
                   border.color = "black",
                   border.size = 0.1,
                   dark.background = F,
                   group.by = "homotypic",
                   cols = c("cyan3","deeppink","white"),
                   flip_xy = FALSE,
                   coord.fixed = TRUE) +
  theme_void() + theme(aspect.ratio = 1) +
  xlim(zoom_fov[, 2]) +
  ylim(zoom_fov[, 1]) + ggtitle(title)

p1 | p2

ggsave("plots/homotypic_patA_fov38.pdf", height=5, width=12)
```


Save object with homo/heterotypic annotation
```{r}
#Clean up a bit the meta-data
to_keep <- c("pat_fov", "orig.ident", "nCount_Nanostring", "nFeature_Nanostring", 
             "cell_ID", "fov", "Area", "AspectRatio", "Width", "Height", "Mean.PanCK", 
             "Max.PanCK", "Mean.CD68", "Max.CD68", "Mean.Membrane", "Max.Membrane", 
             "Mean.CD45", "Max.CD45", "Mean.DAPI", "Max.DAPI", "Slide_name", 
             "Run_name", "Run_Slide_name", 
             "slide_ID_numeric", "Run_Tissue_name.x", "log10totalcounts", "id", "Run_Tissue_name.y", 
             "FOV", "Patient_ID", "Area_ID", "Condition", "Ulcerated_area", 
             "Dist_from_wound", "Wound_direction", "Morphology",
             "prop_neg", "low_quality", "celltype", "celltype_prob",
             "Run_Tissue_name", "Condition2", "MP1.CellCyc", "MP2.RadMig", 
             "MP3.EpiDiff", "MP4.Angiogen", "MP5.Hypoxia", "MP6.Stress", "MP7.WoundResp", 
             "CAF_wound", "CAF_unwound.baseline", "CAF_wound-unwound", "invasiveness", 
             "invasiveCAF", "homotypic", "x", "y", "neighbors_vector", "neighbors_encoding")
seu.obj@meta.data <- seu.obj@meta.data[,to_keep]

saveRDS(seu.obj, "cache/CosMx_BCC_Yerly_2024_wMeta.rds")
```


#Heterotypic/homotypic ratio as a metric of morphological complexity
```{r}
library(dplyr)

tab <- table(seu.obj$pat_fov, seu.obj$homotypic)
tab <- as.data.frame.matrix(tab)
colnames(tab) <- c("homotypic","heterotypic","other")

tab$hetero_fraction <- tab$heterotypic / tab$homotypic
order <- rownames(tab)

#Add metadata
meta <- seu.obj@meta.data[,c("pat_fov","Condition","Condition2","Ulcerated_area",
                             "Dist_from_wound","Morphology")] %>% distinct()
rownames(meta) <- meta$pat_fov
meta <- meta[order,]

tab <- cbind(tab, meta)

#restrict to minimum # of cancer cells
min.cancer.cells <- 200
tab <- tab[tab$homotypic + tab$heterotypic > min.cancer.cells,]

tab$Condition2 <- factor(tab$Condition2, levels=c("Baseline","Unwound","Wound_Far","Wound_Close"))

```

Evaluate heterotypic/homotypic ratio in nodular vs. infiltrative FOVs
```{r fig.height=3.5, fig.width=3}
library(ggpubr)
tab.use <- tab[tab$Condition %in% c("Ulcerated_nodular") & tab$Morphology %in% c("Nodular","Infiltrative"),]

px <- ggplot(tab.use, aes(x=Morphology, y=hetero_fraction, fill=Morphology)) + 
  geom_boxplot(outlier.shape = NA) + theme_classic2() +
  geom_point(aes(fill = Morphology), size = 0.8, shape = 21, position = position_jitterdodge(jitter.width = 0.2)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1), axis.title.x = element_blank()) +
  ggtitle("Morphological complexity") +
  scale_fill_manual(values=c("#F44400","#FFFF6C")) +
  stat_compare_means(label="p.format", size=3) +
  ylim(c(0,2))
px 

ggsave("plots/morph_complexity_by_nodinf.pdf", height=3.5, width=3)
```

Evaluate heterotypic/homotypic ratio in baseline vs. unwound vs. wound (close and far from the wound)
```{r fig.height=5, fig.width=4}
tab.use <- tab[tab$Condition %in% c("Baseline","Unwound","Wound"),]

my_comparisons <- list( c("Baseline", "Wound_Close"), c("Baseline", "Wound_Far"), 
                        c("Unwound", "Wound_Close"))

px <- ggplot(tab.use, aes(x=Condition2, y=hetero_fraction, fill=Condition2)) + 
  geom_boxplot(outlier.shape = NA) + theme_classic2() +
  geom_point(aes(fill = Condition2), size = 0.8, shape = 21, position = position_jitterdodge(jitter.width = 0.2)) +
  theme(axis.text.x = element_text(angle = 30, vjust = 1, hjust=1), axis.title.x = element_blank()) +
  ggtitle("Morphological complexity") +
  scale_fill_manual(values=c("#07f495","#f4ec07","orange","#f44007")) +
  stat_compare_means(comparisons = my_comparisons)
px 

ggsave("plots/morph_complexity_by_wounding.pdf", height=4.5, width=3.5)
```

# Represent interactome within a sample as a graph
```{r}
set.seed(1234)
meta.all <- seu.obj[[]]

base <- meta.all[meta.all$Condition == "Baseline",]
unwound <- meta.all[meta.all$Condition == "Unwound",]
wound <- meta.all[meta.all$Condition == "Wound",]

closest.b <- closest_celltype(base, radius = 100*radius_avg, k=3)
closest.u <- closest_celltype(unwound, radius = 100*radius_avg, k=3)
closest.w <- closest_celltype(wound, radius = 100*radius_avg, k=3)

#kk  layout
make.graph.plot(base, knn.res=closest.b, min.value=3, width.wgt=0.2, layout="kk")
make.graph.plot(unwound, knn.res=closest.u, min.value=3, width.wgt=0.2, layout="kk")
make.graph.plot(wound, knn.res=closest.w, min.value=3, width.wgt=0.2, layout="kk")
```
