---
title: "Correlating MP activity with proximity to the wound in CosMx BCC dataset"
Author: "Massimo Andreatta"
date: "2024-06-18"
editor_options: 
  markdown: 
    wrap: 72
---

Using a CosMx SMI spatial transcriptomics dataset of human BCC, we define an axis to quantify distance and orientation towards the wounding site. For each FOV, we quantify whether MP activity and CAF signature scores correlate with proximity to the wound.

```{r}
library(Seurat)
library(sf)
library(ggplot2)
library(future)
library(RColorBrewer)
library(igraph)
plan("sequential")
```


#Define functions to be used in this workbook
1. Convert cardianal coordinates to angles (West is 0)
```{r}
dir_to_angle <- function(dir, radiants=FALSE) {
  if (dir == "E") {
    angle = 180
  } else if (dir == "SE") {
    angle = 225
  } else if (dir == "S") {
    angle = 270
  } else if (dir == "SW") {
    angle = 315
  } else if (dir == "W") {
    angle = 0
  } else if (dir == "NW") {
    angle = 45
  } else if (dir == "N") {
    angle = 90
  } else if (dir == "NE") {
    angle = 135
  }
  if (radiants) {
    angle <- angle * pi / 180
  }  
  return(angle)
}
```

2. Determine distance from wound for a given FOV, in a consistent system of coordinates across FOVs
```{r}
wound_direction <- function(obj, slide, fov) {
  
  use_slide_metadata <- grep(slide, unique(obj$Run_Tissue_name.x), value=T)
  
  cells_use <- seu.obj$id[(obj$fov == fov) &
                          (obj$Run_Tissue_name.x == use_slide_metadata) &
                          !is.na(obj$x)]
  
  dir <- unique(obj@meta.data[cells_use,"Wound_direction"])
  angle <- dir_to_angle(dir)

  meta.this <- suppressWarnings(subset(seu.obj, cells=cells_use))[[]]
  rad <- angle * pi / 180
  meta.this$xrotate <- meta.this$x * cos(rad) - meta.this$y * sin(rad)
  meta.this$yrotate <- meta.this$x * sin(rad) + meta.this$y * cos(rad)
  
  #recenter
  minx <- min(meta.this$xrotate)
  miny <- min(meta.this$yrotate)
  meta.this$xc <- meta.this$xrotate - minx
  meta.this$yc <- meta.this$yrotate - miny
  meta.this
}
```

#Read in pre-processed data
Pre-process data first, for cell type annotation, with 1-celltype_insitutype.Rmd
```{r}
seu.obj <- readRDS("cache/cosmx_insitutype_anno.rds")
seu.obj$annotation_broad <- seu.obj$celltype
palette <- seu.obj@misc$palette

meta.names <- colnames(seu.obj[[]])
meta.names <- gsub(".x$", "", meta.names, perl=T)
colnames(seu.obj[[]]) <- meta.names
```

Include information for FOV distance from the wound
```{r}
seu.obj$Condition2 <- seu.obj$Condition
ind <- which(!is.na(seu.obj$Dist_from_wound))
seu.obj$Condition2[ind] <- paste(seu.obj$Condition[ind], seu.obj$Dist_from_wound[ind], sep="_")
table(seu.obj$Condition2)
```


#Score MP signatures

Read in cancer MP signatures, and fibroblast signatures
```{r}
mp_file <- "_aux/TableS2_MPs_filtered.csv"

mps <- read.csv(mp_file)
mps <- as.list(mps)

#We can only evaluate genes in the CosMx panel
panel <- rownames(seu.obj)

mps <- lapply(mps, function(x) {
  x[x %in% panel]
})

fb_file <- "_aux/TableS3_wrCAFsignatures.csv"

fb <- read.csv(fb_file)
fb <- as.list(fb)
fb <- lapply(fb, function(x) {
  x[x %in% panel]
})

```

Score meta-programs
```{r}
library(UCell)
seu.obj <- AddModuleScore_UCell(seu.obj, features=c(mps, fb), name="", ncores=8, maxRank = 200)
```

Also calculate differential signatures: MP7 minus MP2, and CAF_wound minus CAF_unwound
```{r}
seu.obj$invasiveness <- seu.obj$MP7.WoundResp - seu.obj$MP2.RadMig
seu.obj$invasiveCAF <- seu.obj$CAF_wound - seu.obj$CAF_unwound.baseline

cancer <- subset(seu.obj, subset=celltype=="Cancer.cells")
caf <- subset(seu.obj, subset=celltype=="Fibroblast")
```

Correlation between MPs in CosMx data
```{r}
library(ggcorrplot)

matrix <- as.matrix(cancer@meta.data[,names(mps)])
colnames(matrix) <- gsub("\\.\\S+$", "", colnames(matrix))
  
corr_matrix <- cor(matrix)
ggcorrplot(corr_matrix, method="circle", type="upper", hc.order = F)

ggsave("plots/MP_scores_correlation_CosMx.pdf", height=3, width=4)
```


Add X and Y coordinates for cell centroids to metadata
```{r}
fovs <- unique(seu.obj$pat_fov)
imgs <- Images(seu.obj)
diet <- DietSeurat(seu.obj)
diet@images <- list()

meta.list <- lapply(fovs, function(f) {
  
  pat <- gsub("\\S+(Patient\\d+)_.*", replacement = "\\1", x = f, perl=T)
  image <- grep(pattern = pat, x = Images(seu.obj), value=T)
  
  this <- subset(diet, subset=pat_fov == f)
  cells_this <- colnames(this)
  
  centroid_data <- seu.obj@images[[image]]$centroids
  centroid_this <- centroid_data@coords[centroid_data@cells %in% cells_this,]
  rownames(centroid_this) <- centroid_data@cells[centroid_data@cells %in% cells_this]
  
  this <- AddMetaData(this, metadata = centroid_this)
  
  this[[]]
})
names(meta.list) <- fovs

meta <- Reduce(f=rbind, x=meta.list)
seu.obj <- AddMetaData(seu.obj, metadata = meta[,c("x","y")])
```


#Can we define an axis of distance from wound?

Define first for one given FOV, to understand the rotation to a common wound axis
```{r fig.width=15, fig.height=4}
use_fov <- 19 # FOV desired 
use_slide_metadata <- unique(seu.obj@meta.data$Run_Tissue_name.x)[1] 
use_slide_image <- Images(seu.obj)[1]

cells_of_interest <- seu.obj$id[(seu.obj$fov == use_fov) & ( seu.obj$Run_Tissue_name.x == use_slide_metadata)] 
centroid_data <- seu.obj@images[[use_slide_image]]$centroids 
zoom_fov <- apply(centroid_data@coords[centroid_data@cells %in% cells_of_interest,], 2, range)

Idents(seu.obj) <- "celltype"
ImageFeaturePlot(seu.obj,
             fov = use_slide_image,
             features = c("MP7.WoundResp"),
             cells = WhichCells(seu.obj, idents = c("Cancer.cells")),
             border.size = 0.2,
             min.cutoff = 0, max.cutoff = 0.3) + xlim(zoom_fov[, 2]) + ylim(zoom_fov[, 1]) +
  coord_flip()
```


```{r fig.height=5, fig.width=10}
meta.this <- wound_direction(seu.obj, fov=use_fov, slide = 1)
meta.cancer <- meta.this[meta.this$celltype=="Cancer.cells",]

a <- ggplot(meta.this, aes(x=xc, y=yc, col=celltype)) + 
  geom_point() + scale_color_manual(values=palette) + 
  theme_light() + theme(aspect.ratio = 1)
b <- ggplot(meta.this, aes(x=x, y=y, col=celltype)) + 
  geom_point() + scale_color_manual(values=palette) +
  theme_light() + theme(aspect.ratio = 1) + NoLegend()
cols <- brewer.pal(9, 'Reds')
c <- ggplot(meta.cancer, aes(x=xc, y=yc, col=MP7.WoundResp))+ 
  geom_point() + theme_light() + theme(aspect.ratio = 1) +
        scale_colour_gradientn(colours = cols)

b | a | c
```

Eval. average signature score by spatial wounding coordinate
```{r}
meta.c <- meta.this[meta.this$celltype == "Cancer.cells",]

ggplot(meta.c, aes(x=xc, y=MP7.WoundResp)) + geom_point() + theme_light()
```

Now we apply this approach systematically on all FOVs
```{r fig.height=15, fig.width=25}
library(patchwork)

fovs_wound <- unique(seu.obj$pat_fov[seu.obj$Condition %in% c("Wound")])

df.list <- lapply(fovs_wound, function(x) {
  fov <- gsub("\\S+_\\S+_(\\d+)$", "\\1", x)
  slide <- gsub("\\S+_(\\S+)_\\d+$", "\\1", x)
  wound_direction(seu.obj, fov=fov, slide = slide)
})
names(df.list) = fovs_wound
```


```{r fig.height=11, fig.width=12}
pll <- lapply(names(df.list), function(n) {
  df <- df.list[[n]]
  df <- df[df$celltype == "Cancer.cells",]
  
  lm <- lm(MP7.WoundResp ~ xc, data=df)
  slope <- ifelse(lm$coefficients[2] >0, yes = "blue", no="red")
  
  ggplot(df, aes(x=xc, y=MP7.WoundResp)) + geom_point(col="gray80", alpha=0.5) +
  geom_smooth(method='lm', formula= y~x, col=slope) +
    ggtitle(n)  + theme_light()
})
wrap_plots(pll)

pll <- lapply(names(df.list), function(n) {
  df <- df.list[[n]]
  df <- df[df$celltype == "Cancer.cells",]
  
  lm <- lm(MP2.RadMig ~ xc, data=df)
  slope <- ifelse(lm$coefficients[2] >0, yes = "blue", no="red")
  
  ggplot(df, aes(x=xc, y=MP2.RadMig)) + geom_point(col="gray80", alpha=0.5) +
  geom_smooth(method='lm', formula= y~x, col=slope) +
    ggtitle(n)  + theme_light()
})
wrap_plots(pll)
```

Summarize by slope
```{r}
cols <- c("red","forestgreen","blue","deeppink","darkcyan","orange","darkmagenta")
names(cols) <- names(mps)

mp.grad <- lapply(names(mps), function(mp) {
  slopes <- lapply(names(df.list), function(n) {
    df <- df.list[[n]]
    df <- df[df$celltype == "Cancer.cells",]
    cor <- cor.test(df[,mp], df[,"xc"], method="spearman")
    unname(-1 * cor$estimate)
  })
  names(slopes) <- names(df.list)
  unlist(slopes)
})
names(mp.grad) <- names(mps)
mp.grad <- as.data.frame(mp.grad)

rownames(mp.grad) <- gsub(".cor","",rownames(mp.grad))
rownames(mp.grad) <- gsub(".rho","",rownames(mp.grad))

avgs <- apply(mp.grad, 2, mean)
avgs <- sort(avgs, decreasing = T)
order <- names(avgs)

mp.grad$Condition <- sapply (rownames(mp.grad), function(p) {
  unique(seu.obj@meta.data[seu.obj$pat_fov == p, "Condition2"])
})

#Now plot distribution by MP
mp.grad <- as.data.frame(mp.grad)
melt <- reshape2::melt(mp.grad)
colnames(melt) <- c("Condition","MP","Correlation")

melt$MP <- factor(melt$MP, levels=order)

pJ <- ggplot(melt, aes(x=MP, y=Correlation, color=MP, shape=Condition)) + 
 geom_jitter(position=position_jitter(0.2)) + theme_classic() +
  scale_color_manual(values=cols) +
   scale_shape_manual(values=c(16, 3)) +
  geom_hline(yintercept=0) +
  theme(axis.text.x = element_text(angle = 30, vjust = 1, hjust=1), axis.title.x = element_blank()) +
  ylab("Correlation with wound proximity") +
  ggtitle("Cancer cells") 
pJ

ggsave("plots/wound_proximity_vs_MP_score.pdf", plot=pJ, height=4, width=6)

```

For CAF signatures
```{r}
caf.sigs <- c(names(fb), "invasiveCAF")

mp.grad <- lapply(caf.sigs, function(mp) {
  slopes <- lapply(names(df.list), function(n) {
    df <- df.list[[n]]
    df <- df[df$celltype == "Fibroblast",]
    cor <- cor.test(df[,mp], df[,"xc"], method="spearman")
    unname(-1 * cor$estimate)

  })
  names(slopes) <- names(df.list)
  unlist(slopes)
})
names(mp.grad) <- caf.sigs
mp.grad <- as.data.frame(mp.grad)

rownames(mp.grad) <- gsub(".cor","",rownames(mp.grad))
rownames(mp.grad) <- gsub(".rho","",rownames(mp.grad))

avgs <- apply(mp.grad, 2, mean)
avgs <- sort(avgs, decreasing = T)
order <- names(avgs)

mp.grad$Condition <- sapply (rownames(mp.grad), function(p) {
  unique(seu.obj@meta.data[seu.obj$pat_fov == p, "Condition2"])
})


#Now plot distribution by MP
mp.grad <- as.data.frame(mp.grad)
melt <- reshape2::melt(mp.grad)
colnames(melt) <- c("Condition","MP","Correlation")

melt$MP <- factor(melt$MP, levels=order)

pK <- ggplot(melt, aes(x=MP, y=Correlation, color=MP, shape=Condition)) + 
 geom_jitter(position=position_jitter(0.2)) + theme_classic() +
  scale_color_manual(values=c("darkred","red","darkblue")) +
  scale_shape_manual(values=c(16, 3)) +
  geom_hline(yintercept=0) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1), axis.title.x = element_blank()) +
  ylab("Correlation with wound proximity") +
  ggtitle("CAF") 
pK

ggsave("plots/wound_proximity_vs_CAF_score.pdf", plot=pK, height=4, width=4)
```



```{r fig.height=5, fig.width=8}
pJ + NoLegend() + pK +  plot_layout(widths = c(2, 1))
ggsave("plots/wound_gradient_vs_MPs.pdf", height=5, width=8)
```
