---
title: "Combination_all_Ganier_BCCs"
author: "Laura Yerly"
output: html_document
date: "2024-02-06"
---

Data are from Ganier et al. (https://www.pnas.org/doi/full/10.1073/pnas.2313326120#data-availability), ArrayExpres: https://www.ebi.ac.uk/biostudies/arrayexpress/studies/E-MTAB-13085

The study contains samples from multiple sites, including 8 samples from BCC:
WS_SKN_KCL9369630, WS_SKN_KCL9369632, WS_SKN_KCL10525740, WS_SKN_KCL10525742, WS_SKN_KCL10525743, WS_SKN_KCL10525745, WS_SKN_KCL10525746, WS_SKN_KCL10525747

```{r setup, include=FALSE,fig.width=16,fig.height=12}
renv::restore()
library(ggplot2)
library(patchwork)
library(Seurat)
library(Matrix)
```

Read count matrices
```{r}
data_dir <- "~/Dropbox/CSI/Collaborations/Kuonen_BCC/data/raw/CellRanger_Ganier"

all_samples <- list.files(path=data_dir, pattern = "^WS_", full.names = T)
sample_ids <- gsub(pattern="\\S+(\\d{3}$)", replacement = "\\1", x=all_samples)
sample_ids <- paste0("CL", sample_ids)
```

Standardize gene symbols while loading
```{r}
library(STACAS)
# load gene symbols dataset
data(EnsemblGeneTable.Hs)

options("Seurat.object.assay.version" = "v3")

seu.list <- lapply(seq_along(all_samples), function(i) {
  
  mtx <- paste0(all_samples[i], "/filtered_feature_bc_matrix")
  
  mat <- Read10X(mtx)
  a <- CreateSeuratObject(mat, min.features = 50, min.cells = 0)
  a <- StandardizeGeneSymbols(a, slots = "counts", EnsemblGeneTable = EnsemblGeneTable.Hs)
  
  a$Sample <- sample_ids[i]
  NormalizeData(a)
})
```

#Merge into single object
```{r}
data_seurat <- Reduce(f=merge, x=seu.list)
#data_seurat <- JoinLayers(data_seurat)
```

## Dissociation-related genes
Dissociation-related genes: Genes associated with the dissociation
protocol, i.e. stress genes, can sometimes cause clustering of cells in
different groups. Described by van den brink et al 2017
<https://www.nature.com/articles/nmeth.4437#accession-codes> Another
paper (Fan et al 2019) used the genes in Van Den Brink et al. (2017) and
posted the Rmd with the code to remove genes on github:
<https://www.nature.com/articles/s41467-019-11036-9#code-availability>
removed cells that expressed high dissociation-related genes. Eg in
Rachel's analysis of Jeremiah's data, she saw that a 10% dissoc-related
gene cut off was removing many cells, one cluster was very high for
those genes, but even regressing-out the genes would not remove fully
this cluster that they couldn't really annotate, so they removed it.
Gene list is from Van Den Brink et al. (2017) R code in supplementary
method under "In silico purification"
```{r}
dissoc_genes <- c("Actg1__chr11","Ankrd1__chr19","Arid5a__chr1","Atf3__chr1","Atf4__chr15","Bag3__chr7","Bhlhe40__chr6",
                  "Brd2__chr17","Btg1__chr10","Btg2__chr1","Ccnl1__chr3","Ccrn4l__chr3","Cebpb__chr2","Cebpd__chr16",
                  "Cebpg__chr7","Csrnp1__chr9","Cxcl1__chr5","Cyr61__chr3","Dcn__chr10","Ddx3x__chrX","Ddx5__chr11",
                  "Des__chr1","Dnaja1__chr4","Dnajb1__chr8","Dnajb4__chr3","Dusp1__chr17","Dusp8__chr7",
                  "Egr1__chr18","Egr2__chr10","Eif1__chr11","Eif5__chr12","Erf__chr7","Errfi1__chr4","Fam132b__chr1",
                  "Fos__chr12","Fosb__chr7","Fosl2__chr5","Gadd45a__chr6","Gcc1__chr6","Gem__chr4","H3f3b__chr11",
                  "Hipk3__chr2","Hsp90aa1__chr12","Hsp90ab1__chr17","Hspa1a__chr17","Hspa1b__chr17","Hspa5__chr2",
                  "Hspa8__chr9","Hspb1__chr5","Hsph1__chr5","Id3__chr4","Idi1__chr13","Ier2__chr8","Ier3__chr17",
                  "Ifrd1__chr12","Il6__chr5","Irf1__chr11","Irf8__chr8","Itpkc__chr7","Jun__chr4","Junb__chr8",
                  "Jund__chr8","Klf2__chr8","Klf4__chr4","Klf6__chr13","Klf9__chr19","Litaf__chr16","Lmna__chr3",
                  "Maff__chr15","Mafk__chr5","Mcl1__chr3","Midn__chr10","Mir22hg__chr11","Mt1__chr8","Mt2__chr8",
                  "Myadm__chr7","Myc__chr15","Myd88__chr9","Nckap5l__chr15","Ncoa7__chr10","Nfkbia__chr12","Nfkbiz__chr16",
                  "Nop58__chr1","Nppc__chr1","Nr4a1__chr15","Odc1__chr12","Osgin1__chr8","Oxnad1__chr14","Pcf11__chr7",
                  "Pde4b__chr4","Per1__chr11","Phlda1__chr10","Pnp__chr14","Pnrc1__chr4","Ppp1cc__chr5","Ppp1r15a__chr7",
                  "Pxdc1__chr13","Rap1b__chr10","Rassf1__chr9","Rhob__chr12","Rhoh__chr5","Ripk1__chr13","Sat1__chrX",
                  "Sbno2__chr10","Sdc4__chr2","Serpine1__chr5","Skil__chr3","Slc10a6__chr5","Slc38a2__chr15",
                  "Slc41a1__chr1","Socs3__chr11","Sqstm1__chr11","Srf__chr17","Srsf5__chr12","Srsf7__chr17",
                  "Stat3__chr11","Tagln2__chr1","Tiparp__chr3","Tnfaip3__chr10","Tnfaip6__chr2","Tpm3__chr3",
                  "Tppp3__chr8","Tra2a__chr6","Tra2b__chr16","Trib1__chr15","Tubb4b__chr2","Tubb6__chr18",
                  "Ubc__chr5","Usp2__chr9","Wac__chr18","Zc3h12a__chr4","Zfand5__chr19","Zfp36__chr7","Zfp36l1__chr12",
                  "Zfp36l2__chr17","Zyx__chr6","Gadd45g__chr13","Hspe1__chr1","Ier5__chr1","Kcne4__chr1")
dissoc_genes <- toupper(sapply(dissoc_genes, function(x){
  strsplit(x, "__")[[1]][1]
}))
length(dissoc_genes) # 140
head(dissoc_genes)
```

```{r define_qc_metrics}
#Get mitochondrial and ribosomal signatures
library(SignatuR)
ribo.genes <- GetSignature(SignatuR$Hs$Compartments$Ribo)
mito.genes <- GetSignature(SignatuR$Hs$Compartments$Mito)


# Compute ribosomal and mitochondrial content and add to Seurat object metadata
data_seurat <- AddMetaData(data_seurat, metadata = PercentageFeatureSet(data_seurat, features = ribo.genes[ribo.genes %in% rownames(data_seurat)]), col.name = "percent.ribo")

data_seurat <- AddMetaData(data_seurat, metadata = PercentageFeatureSet(data_seurat, features = mito.genes[mito.genes %in% rownames(data_seurat)]), col.name = "percent.mito")

data_seurat <- AddMetaData(data_seurat, metadata = PercentageFeatureSet(data_seurat, features = dissoc_genes[dissoc_genes %in% rownames(data_seurat)]), col.name = "percent.dissoc")

# Compute the ratio of number of genes/features and number of counts/UMIs
data_seurat$log10GenesPerUmi <- log10(data_seurat$nFeature_RNA) / log10(data_seurat$nCount_RNA)

```




```{r qc_plots0, fig.height= 10}
Idents(data_seurat) <- "Sample"
VlnPlot(data_seurat,
        features = c("nFeature_RNA", "nCount_RNA",
                     "percent.ribo","percent.mito",
                     "percent.dissoc","log10GenesPerUmi"),
        ncol = 3, pt.size=0)
```


### Custom thresholds

```{r set_cutoffs}
cutoffs <- list()
cutoffs[["percent.ribo"]] <- c(min=0,max=60)
cutoffs[["percent.mito"]] <- c(min=0,max=20)
cutoffs[["nFeature_RNA"]] <- c(min=500,max=6000) 
cutoffs[["nCount_RNA"]] <- c(min=600,max=25000)
cutoffs[["percent.dissoc"]] <- c(min=0, max=20)
cutoffs[["log10GenesPerUmi"]] <- 0.8
print(cutoffs)
```

### Standard check

```{r probs}
for(va in names(cutoffs)){
  cat(va, "\n")
  print(quantile(data_seurat@meta.data[[va]],
                 probs=c(0.001, 0.005, 0.01, 0.02, 0.1, 0.50, 0.9,
                         0.98, 0.99, 0.995, 0.999))
        )
}
```



```{r subset_seurat fig.height= 10}
# store initial number of cells
initialCellNum <- ncol(data_seurat)
message(paste("Original number of cells:", initialCellNum))

# Perform subsetting using cutoffs list and qc metrics considered
data_seurat <- subset(data_seurat, subset = 
                        nFeature_RNA >= cutoffs[["nFeature_RNA"]]["min"] &
                        nFeature_RNA < cutoffs[["nFeature_RNA"]]["max"] & 
                        nCount_RNA   >= cutoffs[["nCount_RNA"]]["min"]   & 
                        nCount_RNA   < cutoffs[["nCount_RNA"]]["max"]   &  
                        percent.ribo >= cutoffs[["percent.ribo"]]["min"] & 
                        percent.ribo < cutoffs[["percent.ribo"]]["max"] &
                        percent.mito >= cutoffs[["percent.mito"]]["min"] &
                        percent.mito < cutoffs[["percent.mito"]]["max"] &
                        percent.dissoc >= cutoffs[["percent.dissoc"]]["min"] &
                        percent.dissoc < cutoffs[["percent.dissoc"]]["max"] &
                        log10GenesPerUmi > cutoffs[["log10GenesPerUmi"]]
                      )

# Print and evaluate the cell drop out after filtering
l <- ncol(data_seurat)
message(sprintf("Number of cells after QC: %i (%.2f %% of input)", l, 100*l/initialCellNum))
```


## Data scaling and dimensionality reduction

```{r}
bcc_8CG <- data_seurat
bcc_8CG <- NormalizeData(bcc_8CG) |> FindVariableFeatures(nfeatures=2000) |>
  ScaleData()

bcc_8CG <- bcc_8CG |> RunPCA(npcs=30) |> RunUMAP(reduction = "pca", dims = 1:30)

a <- DimPlot(bcc_8CG, reduction = "umap", group.by = "Sample") + theme(aspect.ratio = 1)
a
```

## Clustering

```{r}
bcc_8CG <- FindNeighbors(object = bcc_8CG, dims=1:30)
bcc_8CG <- FindClusters(object = bcc_8CG, resolution = 1)
DimPlot(bcc_8CG, reduction = "umap", group.by = "seurat_clusters") + theme(aspect.ratio = 1)
```

##Broad classification using scGate

```{r eval=T}
library(scGate)

models.DB <- scGate::get_scGateDB(destination = ".", version = "v0.16")
models.list <- models.DB$human$TME_HiRes

models.list$Melanocyte <- gating_model(name = "Melanocyte", signature = c("PMEL","MLANA","LYZ-"))

bcc_8CG <- scGate(bcc_8CG, model = models.list, ncores = 8, multi.asNA = T)
```

```{r}
b <- DimPlot(bcc_8CG, group.by = "scGate_multi", label = T, repel = T, label.size = 2) +
    theme(aspect.ratio = 1) + ggtitle("scGate annotation")
b
```

```{r fig.width=12}
a | b
```


```{r}
table(bcc_8CG$scGate_multi, useNA = "ifany")
table(bcc_8CG$scGate_multi, bcc_8CG$Sample, useNA = "ifany")
```


## Save 
```{r}
file <- "cache/bcc_8CG.rds"

saveRDS(bcc_8CG, file)
```
