---
title: "Apply NMF on BCC samples"
author: "Massimo Andreatta and Laura Yerly"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---


```{r, include=FALSE, fig.width=16, fig.height=12}
renv::restore()
library(Seurat)
library(ggplot2)
library(SignatuR)
library(UCell)
library(patchwork)
library(tidyr)
library(dplyr)
library(RColorBrewer)

#install.packages("GeneNMF")
#remotes::install_github("carmonalab/GeneNMF", ref="dev")
library(GeneNMF)
```


## Set the paths

```{r}
data.path <- '~/Dropbox/CSI/Collaborations/Kuonen_BCC/data/rds/Tumor_combination_LY_CG.rds'
#data.path <- "cache/Tumor_combination_LY_CG.rds"
seu <- readRDS(data.path)
```


Prep data
```{r}
DefaultAssay(seu) <- "RNA"

seu.list <- SplitObject(seu, split.by = "Sample")
```


##Using GeneNMF version 0.4.0 (from CRAN)
```{r message=F, results=F, warning=F, echo=FALSE}
set.seed(123)

geneNMF.programs <- multiNMF(seu.list, assay="RNA", slot="data", k=4:9,
                    min.exp = 0.05, max.exp = 3, nfeatures = 2000, seed=123)
```

Extract meta-programs
```{r}
geneNMF.metaprograms <- getMetaPrograms(geneNMF.programs, max.genes=200,
                                        hclust.method="ward.D2", nprograms=10,
                                        min.confidence=0.3)
```

MP genes:
```{r}
a <- lapply(geneNMF.metaprograms$metaprograms.genes, function(x){head(x, n=20)})
a
```

MP statistics:
```{r}
geneNMF.metaprograms$metaprograms.metrics
```


Heatmap of MP similarity
```{r fig.width=10, fig.height=8}
ph <- plotMetaPrograms(geneNMF.metaprograms, jaccard.cutoff = c(0,1))

ggsave("plots/BCC_heatmap_jaccard.pdf", plot=ph, height=7, width=9)
```

Set criteria to remove meta-programs:
```{r}
geneNMF.metaprograms$metaprograms.metrics

tokeep <- geneNMF.metaprograms$metaprograms.metrics[geneNMF.metaprograms$metaprograms.metrics$silhouette > 0 &
                                                        geneNMF.metaprograms$metaprograms.metrics$meanJaccard > 0.01 &
                                                        geneNMF.metaprograms$metaprograms.metrics$numberGenes >= 5, ]

geneNMF.metaprograms$metaprograms.metrics <- tokeep

geneNMF.metaprograms$metaprograms.genes <- geneNMF.metaprograms$metaprograms.genes[rownames(tokeep)]
```

NOTE: MP6 and MP8 are largely overlapped. These could be merged or one could be discarded. Because MP6 one has a significantly higher sample coverage, we can discard the other one.

Quantify similarity between MPs:

```{r}
overlap_coef <- function(vec1, vec2) {
  intersection <- length(intersect(vec1, vec2))
  min_size <- min(c(length(vec1), length(vec2)))
  return(intersection / min_size)
}


num_vectors <- length(geneNMF.metaprograms$metaprograms.genes)
similarity_matrix <- matrix(NA, nrow = num_vectors, ncol = num_vectors)
max_overlap_coef <- rep(0,length(geneNMF.metaprograms$metaprograms.genes))

# Compute similarity between all pairs of vectors
for (i in 1:num_vectors) {
  for (j in 1:num_vectors) {
    if (i != j) {
      similarity_matrix[i, j] <- overlap_coef(geneNMF.metaprograms$metaprograms.genes[[i]], geneNMF.metaprograms$metaprograms.genes[[j]])
      if (similarity_matrix[i, j] > max_overlap_coef[i]) { max_overlap_coef[i] <- similarity_matrix[i, j]}  
    } else {
      similarity_matrix[i, j] <- 1  
    }
  }
}
rownames(similarity_matrix) <- rownames(geneNMF.metaprograms$metaprograms.metrics)
colnames(similarity_matrix) <- rownames(geneNMF.metaprograms$metaprograms.metrics)

print(similarity_matrix)
heatmap(similarity_matrix)
```

```{r}
geneNMF.metaprograms$metaprograms.metrics$maxOverlapCoef <- max_overlap_coef
geneNMF.metaprograms$metaprograms.metrics
```



```{r}
tokeep <- geneNMF.metaprograms$metaprograms.metrics[rownames(geneNMF.metaprograms$metaprograms.metrics) != "MP9",] # large gene overlap (>40%) with MP6 and low sampleCoverage (<40%)
geneNMF.metaprograms$metaprograms.metrics <- tokeep
geneNMF.metaprograms$metaprograms.genes <- geneNMF.metaprograms$metaprograms.genes[rownames(tokeep)]
rownames(tokeep)
```

```{r}
#Save gene list to file
as.df <- t(plyr::ldply(geneNMF.metaprograms$metaprograms.genes, rbind))
write.csv(as.df, file = "_aux/NMF.genes.10mp.csv")
```

# Save results
```{r}
saveRDS(geneNMF.metaprograms, file="cache/NMF_meta_res.rds")
```

