---
title: "Study the effect of surgery on BCC tumor microenvironment"
author: "Josep Garnica"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

The goal of this script is to study the tumor microenvironment of the wounding longitudinal experiment. Original data can be accessed from The data can be downloaded from GEO (GSE266665): https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE266665. Processing of the data as well as annotation is found in scripts 4*.Rmd.

Load needed packages
```{r, include=FALSE, fig.width=16, fig.height=12}
renv::restore()
library(Seurat)
library(scGate)
library(ggplot2)
library(UCell)
library(tidyr)
library(dplyr)
library(tibble)

```


# Load data
Read in BCC longitudinal experiment:
Two patients at three timepoints: Baseline (first biopsy), 1 week after surgery on same wound, 1 week after surgery at different location. 

Original data can be accessed from The data can be downloaded from GEO (GSE266665): https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE266665. 
Processing of the data is found in scripts 4*.Rmd.

```{r}
file_wh <- "cache/WHexp_annotated.rds"
file_wh <- "~/Dropbox/CSI/Collaborations/Kuonen_BCC/data/rds/WHexp_annotated.rds"

p <- readRDS(file_wh)
```


# Annotate with scGate
Load scGate models
```{r, warning=F}
# use SHA of the models for reproducibility 
models <- get_scGateDB(branch = "1f650e5bfec2adb6ba121dcf90b6b0aef964c7ea",
                       force_update = T)
scgate.model <- models$human$HiTME

```

```{r}
# split object by sample into a list
spl <- SplitObject(p,
                   split.by = "patient_bcc")

# Run scGate models independently for each sample
spl <- lapply(spl,
              function(s){
                scGate(s,
                       model = scgate.model,
                       multi.asNA = T,
                       ncores = 9)
              })
```



# Fine annotation of Stromal/Fibroblasts cells
## Keep only Fibroblasts/stromal cells, removing cancer cells

```{r}

seu.fibro <- lapply(spl,
                    function(s){
                      subset(s, scGate_multi == "Fibroblast" &
                               annotation != "Cancer_cells")
                    })
```


## Load scGate models for Fibroblasts, pericytes and smooth muscle cells (SMC)
Cancer-associated fibroblasts (CAF) were identified using the scGate method with specific fibroblast markers, excluding pericytes (PDGFA+, FBLN1+, FBLN2+, COL5A1+, LUM+, CFD+, RGS5-, NOTCH3-, MYOT-) and aggregated into a pseudo-bulk expression matrix for each sample. 

```{r}
pat <- "_scGate.tsv"
dir <- "_aux/"

mods.file <- list.files(dir, pattern = pat)

mods <- lapply(mods.file, 
               function(f){
                 read.table(file.path(dir, f),
                            header = T)
               })

names(mods) <- gsub(pat, "", mods.file)
```

## Fine annotation only Fibroblasts/stromal cells into bona fide fibroblasts, pericytes and SMC
```{r}

seu.fibro <- lapply(seu.fibro,
                    function(s){
                      scGate::scGate(s,
                                     min.cells = 15,
                                     model = mods,
                                     ncores = 6,
                                     pca.dim = 20)
                    })
```


# TME proportions

## Ratio of ratios:
Do ratio of most relevant cell types in wounded samples,
Compute the ratio of relative abundances of tumor microenvironment (TME) cell types (annotated by scGate) with respect to cancer cells, calculated between wounded and baseline paired samples (log2 scale, N=2). Bars represent the mean ratio for each cell type.

```{r}
# combine scGate annotation and cancer cells manual annotation
pmeta <- lapply(spl, function(s){
  m <- s@meta.data[, c("patient_bcc", "annotation", "scGate_multi")] %>% 
    mutate(category = substr(patient_bcc, 1, nchar(patient_bcc)-1),
           dataset = substr(patient_bcc, nchar(patient_bcc), nchar(patient_bcc)) 
    ) %>% 
    rownames_to_column("rn")
  return(m)
}) %>% 
  data.table::rbindlist(fill = T)


## fibroblast fine annotation
fine <- lapply(seu.fibro, function(s){
  m <- s@meta.data[,"scGate_multi", drop = F]%>% 
    rownames_to_column("rn") %>% 
    mutate(scGate_multi = ifelse(is.na(scGate_multi),
                                 "unknown", scGate_multi))
  names(m)[2] <- "fine_fibroblast"
  return(m)
}) %>% 
  data.table::rbindlist(fill = T)

# join whole annotation and fine fibroblast annotation
cu <- left_join(pmeta, fine, by = "rn")

cts <- c("CD4T", "CD8T", "MoMac", "panDC", "Fibroblast",
         "Pericyte", "Cancer_cells", "Endothelial")
cts2 <- c("CD4T", "CD8T", "MoMac", "panDC", "CAF",
         "Pericyte", "Cancer_cells", "Endothelial")

cu2 <- cu %>% 
  mutate(annot = ifelse(annotation %in% c("Cancer_cells", "Doublets"),
                        annotation,
                        ifelse(!is.na(fine_fibroblast),
                               fine_fibroblast,
                               as.character(scGate_multi)
                        )
  )
  ) %>% 
  filter(annot %in% cts) %>% 
  mutate(annot = factor(annot,
                        levels = cts,
                        labels = cts2)) %>% 
  # keep 0s when counting
  dplyr::count(patient_bcc, annot, .drop = F) %>% 
  mutate(count = n+1, # apply pseudocount for conditions with cell types with 0 cells
         ratio = count/count[annot == "Cancer_cells"], # do ratio vs Cancer cells
         r2 = log2(ratio),
         category = substr(patient_bcc, 1, nchar(patient_bcc)-1), # name of the condition
         category2 = ifelse(category == "Wound",
                            "Wound", "UnWound"),
         dataset = substr(patient_bcc, nchar(patient_bcc), nchar(patient_bcc))) %>% 
  filter(annot != "Cancer_cells") %>% 
  # Filter unwound samples
  filter(category != "Unwound")


rcu <- cu2 %>% 
  group_by(annot, category2, dataset) %>% 
  summarize(ratio_mean = mean(ratio)) %>% 
  group_by(annot, dataset) %>% 
  summarize(rr = ratio_mean[category2 == "Wound"] / 
              ratio_mean[category2 == "UnWound"],
            ratio = log2(rr))

```

### Barplots

```{r}
# ratio plots
plp <- rcu %>% 
  ggplot(aes(annot, ratio)) +
  geom_bar(stat = "summary",
           fun = "mean",
           fill = "grey86",
           color = "black") +
  geom_point(show.legend = F,
             size = 3,
             shape = 21,
             fill = "grey15",
             color = "black") +
  ggprism::theme_prism() +
  labs(title = "TME composition",
       y = "log2 ratio Wound vs Baseline",
       x = "Cell type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+
  NoLegend()

plp

```
### Lineplots
Same information but pairing the data from each patient:

```{r}
# line plots
lycols <- c("Baseline" = "#DEEBF7",
            "Unwound" = "#9ECAE1",
            "Wound" = "#3182BD")
plp <- cu2 %>% 
  ggplot(aes(category, r2, fill = category)) +
  geom_line(aes(group = dataset),
            show.legend = F,
            color = "black") +
  geom_point(aes(group = category),
             size = 3,
             shape = 21,
             color = "black") +
  scale_fill_manual(values = lycols) +
  facet_wrap(~annot, strip.position = "top") +
  ggprism::theme_prism() +
  labs(title = "TME composition",
       y = "log2 ratio vs cancer cells",
       x = "") +
  theme(
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    legend.text = element_text(size = 12)) +
  guides(fill = guide_legend(override.aes = list(size=4)))

plp

```

# Differential expression of TME pools
Let's perform differential expression of pseudobulk for each sample, comparing Wound vs. unwound+baseline.

## Pseudobulk samples

First do pseudobulk of all cell types in `cts`, for which we show proportions earlier.

```{r}
ct = cts[!cts %in% c("Cancer_cells", "Fibroblast", "Pericyte")]

pseudo.p <- list()

for(c in ct){
  mat <- lapply(spl,
                function(s){
                  s@meta.data <- s@meta.data %>% 
                    mutate(annot = ifelse(annotation %in% c("Cancer_cells", "Doublets"),
                                          annotation, scGate_multi)
                    )
                  subs <- s@meta.data %>% filter(annot == c)
                  # keep only cell type with at least 1 cell
                  if(nrow(subs)>1){
                    m <- AggregateExpression(s[,s$annot == c],
                                             group.by = "patient_bcc")[["RNA"]] %>%
                      as.matrix() %>% 
                      t() %>%
                      as.data.frame() %>%
                      rownames_to_column("sample")
                    m$sample <- unique(s$patient_bcc)
                    return(m) 
                  } else {
                    return(NULL)
                  }
                }) %>%
    # bind resulting dataframes
    data.table::rbindlist(fill = T, use.names = T) %>%
    column_to_rownames("sample") %>%
    # convert NAs to 0
    mutate_if(is.numeric, ~ifelse(is.na(.), 0, .)) %>%
    t()
  
  pseudo.p[[c]] <- mat
}

```
Extract also pseudobulk for Fibroblast and pericytes from finely annotated object

```{r}
ct = c("Fibroblast", "Pericyte")
for(c in ct){
  mat <- lapply(seu.fibro,
                function(s){
                  subs <- s@meta.data[s@meta.data$scGate_multi == c,]
                  # keep only cell type with at least 1 cell
                  if(nrow(subs)>1){
                    m <- AggregateExpression(s[,s$scGate_multi == c],
                                             group.by = "patient_bcc")[["RNA"]] %>%
                      as.matrix() %>% 
                      t() %>%
                      as.data.frame() %>%
                      rownames_to_column("sample")
                    m$sample <- unique(s$patient_bcc)
                    return(m) 
                  } else {
                    return(NULL)
                  }
                }) %>%
    # bind resulting dataframes
    data.table::rbindlist(fill = T, use.names = T) %>%
    column_to_rownames("sample") %>%
    # convert NAs to 0
    mutate_if(is.numeric, ~ifelse(is.na(.), 0, .)) %>%
    t()
  
  pseudo.p[[c]] <- mat
  
}
```

## Differential expression
Let's use pseudobulk matrices to perform the differential expression analysis with DESeq2.
Paired differential expression analysis was performed using the DESeq2 package63 on pseudo-bulk samples, comparing wounded vs. unwounded and baseline conditions.

```{r}
require(DESeq2)

res.list <- list()

for(c in names(pseudo.p)){
  
  mat <- pseudo.p[[c]]
  
  md.agg <- data.frame(sample = colnames(mat)) %>% 
    mutate(patient_bcc = sample) %>% 
    mutate(category = substr(patient_bcc, 1, nchar(patient_bcc)-1),
           category2 = ifelse(category == "Wound",
                              "Wound", "Unwound"),
           dataset = substr(patient_bcc, nchar(patient_bcc), nchar(patient_bcc))
    ) %>% 
    column_to_rownames("sample")
  
  
  dds <- DESeq2::DESeqDataSetFromMatrix(countData = mat,
                                        colData = md.agg,
                                        # indicate the pairing
                                        design = ~ dataset + category2) 
  #Remove genes with very low counts:
  dds <- dds[rowSums(counts(dds)) > 6, ]
  
  # do Differential expression
  de <- DESeq2::DESeq(dds)
  
  res <- DESeq2::lfcShrink(de,
                           coef = DESeq2::resultsNames(de)[3],
                           type = "apeglm")
  
  res.list[[c]] <- as.data.frame(res) 
  
}

```

## Fold change plot

This plot show only differential expressed genes, without showing the p-value.
It shows the number of differentially expressed genes (adjusted p-value < 0.05 and |log2 fold-change| > 0.5; DEG), and log2 fold-change of the DEG in 7 major components of the TME.

```{r}
col2 <- cols <-  c("Unwound" = "firebrick4",
                   "Wound" = "blue4")

# accommodate cell type names
le <- c("CD4T", "CD8T", "MoMac", "panDC", "Fibroblast",
        "Pericyte", "Endothelial")
la <- c("CD4T", "CD8T", "MoMac", "panDC", "CAF",
        "Pericyte", "Endothelial")


li2 <- lapply(names(res.list), 
              function(v){
                res.list[[v]] %>%
                  rownames_to_column("gene") %>% 
                  mutate(celltype = v) %>% 
                  filter(padj < 0.05) %>% 
                  filter(abs(log2FoldChange) > 0.5)
              }) %>% 
  data.table::rbindlist() %>% 
  mutate(category = ifelse(sign(log2FoldChange)>0,
                           "Wound", "Unwound"),
         celltype = factor(celltype,
                           levels = le,
                           labels = la),
         sign = sign(log2FoldChange)
  )

nde <- table(li2$celltype) %>% sort()

li2 <- li2 %>% 
  mutate(celltype = factor(celltype,
                           levels = names(nde)))

fc <- 0.5        
set.seed(22)
hb <- ggplot(li2, aes(x=log2FoldChange, y=celltype,
                      color=category,
                      label= gene
)) +
  geom_jitter(size = 2, 
              height=0.15,
              width = 0) +
  scale_color_manual(values = col2)+
  geom_vline(xintercept = c(-fc, fc),
             linetype = 2,
             size = 0.3,
             col = "grey20") +
  scale_y_discrete(drop = F) +
  xlim(c(-7,7)) +
  labs(y = "Cell type",
       x = "log2 Fold Change")+
  ggprism::theme_prism()

# Add text showing up and down genes
lol <- table(li2$log2FoldChange>0, li2$celltype) %>%
  as.data.frame() %>% filter(Var1 == TRUE)
lol$Var1 <- "Wound"
lol2 <- table(li2$log2FoldChange<0, li2$celltype) %>%
  as.data.frame() %>% filter(Var1 == TRUE)
lol2$Var1 <- "Unwound"
lol3 <- rbind(lol2, lol) 

names(lol3) <- c("category", "celltype", "Freq")

lol3 <- lol3 %>% 
  mutate(celltype = factor(celltype,
                           levels = names(nde))) %>% 
  arrange(celltype) %>% 
  mutate(x = rep(c(-6.6,6.6), 7),
         y =  c(1,1,2,2,3,3,4,4,5,5,6,6,7,7)
  )




hb2 <- hb + geom_text(
  data    = lol3,
  mapping = aes(x = x, y = y,
                color = category,
                label = as.character(Freq),
                fontface=2),
  size= 5,
  show.legend = F
)

hb2
```


## Volcano plot for fibroblasts
Show volcano plot for CAF using DE previously obtained.
Differentially expressed genes in CAFs from wounded (N=2) vs. baseline (N=2) and unwounded (N=2) samples. Y axis shows -log10 of adjusted p-values, with a threshold at adj. p-val = 0.05. X axis shows log2 expression fold-change, with thresholds at |log2 fold-change| > 0.5. 

Built function to produce the Volcano

```{r}
require(scales)
volpl <- function(df, fc=2, fdr=0.05, tit = NA,
                  NS = "NS", UP = "UP", DOWN = "DOWN",
                  legend.title = "Differential expression"){
  
  cols <-  c("firebrick4","grey", "blue4")
  names(cols) <- c(DOWN, NS, UP)
  
  Vol <- data.frame(log2FC= df$log2FoldChange,
                    sig= -log10(df$padj), 
                    gene = rownames(df))
  Vol$S <- 0
  Vol[which(Vol$sig<(-log10(fdr))), "S"] <- NS
  Vol[which(Vol$log2FC> fc & Vol$S!="NS"), "S"] <- UP
  Vol[which(Vol$log2FC < (-fc) & Vol$S!="NS"), "S"] <- DOWN
  Vol[which(Vol$S==0), "S"] <- NS
  Vol$S <- as.factor(Vol$S)
  Vol$sig[Vol$sig == "Inf"] <- 320
  Vol <- Vol[complete.cases(Vol),]
  v_5 <- ggplot(Vol, aes(x=log2FC, y=sig,
                         color = S, label=gene)) +
    geom_point (size = 2, alpha=0.7,
                show.legend = T) +
    geom_vline(xintercept = c(-fc, fc), linetype = 1, size = 0.3, col = "grey20") +
    geom_hline(yintercept = -log10(fdr), linetype = 1, size = 0.3, col = "grey20") +
    scale_color_manual(values = cols) +
    ggrepel::geom_text_repel(data = subset(Vol, (sig >= (-log10(fdr)) & abs(log2FC)> fc)),
                             aes(label = Vol[(Vol$sig >= (-log10(fdr)) &
                                               abs(Vol$log2FC) > fc),"gene"]),
                             color="black",
                             size=3,
                             force = 30,
                             max.overlaps = 30)+
    labs(x = "log2 Fold Change",
         y = "-log10 adj.p-value",
         color = legend.title,
         title = tit)+
    scale_y_continuous(trans = pseudo_log_trans())+
    scale_x_continuous(trans = pseudo_log_trans())+
    ggprism::theme_prism() +
    theme(
      legend.text = element_text(size = 12)) +
    guides(fill = guide_legend(override.aes = list(size=4)))
  
  return(v_5)        
}
```


```{r}
volpl(res.list$Fibroblast,
      fc = 0.5,
      tit = "CAF Wound vs Unwound")
```


## Obtain CAF wounding signatures
Genes with a log2 fold change > 0.5 and adjusted p-value < 0.05 were identified as part of the wound signature, while those with a log2 fold change < -0.5 and adjusted p-value < 0.05 constituted the unwound/baseline signature.

```{r}
res0 <- res.list$Fibroblast %>% 
  filter(padj<0.05 & abs(log2FoldChange) > 0.5) %>% 
  mutate(type = ifelse(sign(log2FoldChange)>0,
                   "Wound", "Unwound")
    )

res.spl <- split(res0, f = res0$type)

lys <- lapply(res.spl, function(l){
  l %>% 
    mutate(fc = abs(log2FoldChange)) %>% 
    rownames_to_column("gene") %>% 
    pull(gene)
})

saveRDS(lys, "cache/CAF_signatures.rds")

```








