---
title: "Characterization of cell populations (Ganier et al. set)"
author: "Massimo Andreatta & Laura Yerly"
output: html_document
date: "2024-02-07"
---

```{r setup, include=FALSE,fig.width=16,fig.height=12}
renv::restore()
library(ggplot2)
library(Seurat)
library(RColorBrewer)
library(patchwork)
library(ggrastr)
```

# Read in pre-processed data

Read scRNA-seq data after QC from 2-Combination... script.
```{r}
data.path <- "cache/bcc_8CG.rds"

seu <- readRDS(data.path)
```

## Set up some helper functions
```{r}
get_means_signature<- function(object,signature){
  means <- sapply(levels(object$seurat_clusters), function(x) {
    mean(object@meta.data[object$seurat_clusters == x, signature])
  })
  means
}

get_medians_signature<- function(object,signature){
  medians <- sapply(levels(object$seurat_clusters), function(x) {
    median(object@meta.data[object$seurat_clusters == x, signature])
  })
  medians
}

palette <- RColorBrewer::brewer.pal(10, name="Paired")
names(palette) <- c("Cancer_cells","Normal_Kerat","Melanocyte","Endothelial",
                                                  "Fibroblast","Mast","MoMacDC","Bcell","PlasmaCell","Tcell")  
```

# Look at the data

```{r fig.width=12}
a <- DimPlot(seu, reduction = "umap", group.by = "Sample") + theme(aspect.ratio = 1)
b <- DimPlot(seu, group.by = "scGate_multi", label = T, repel = T, label.size = 2) +
    theme(aspect.ratio = 1) + ggtitle("scGate annotation")
a | b
```

Simplify annotations
```{r}
seu$scGate_broad <- seu$scGate_multi

seu$scGate_broad[seu$scGate_multi %in% c("Macrophage","Monocyte","panDC")] <- "MoMacDC"
seu$scGate_broad[seu$scGate_multi %in% c("CD4T","CD8T","NK")] <- "Tcell"
```

## Signatures to identify cell populations 

```{r}
library(UCell)
signatures<-list()
signatures$BCC<-c("PTCH1","GLI1","GLI2","HHIP","MYCN","SPON2")
signatures$TCells<-c("CD2","CD3D","CD3E")
signatures$NormKerat<-c("KRT14","KRT5","GATA3","KRTDAP")
signatures$Melanocytes<-c("PMEL","MLANA")
signatures$Cycling<-c("MKI67","ORC1")
signatures$MyeloidCells<-c("LYZ","SPI1")
signatures$BCells<-c("CD79A","MS4A1")
signatures$Fibroblasts<-c("COL1A1","IL6","PDGFRA","CXCL12","CD82","CCL19")
signatures$Pericytes<-c("RGS5","KCNE4","CPE")
signatures$MastCells<-c("CTSG","TPSB2","TPSAB1")
signatures$EndothelialCells<-c("PECAM1","HEY1","CCL14","ACKR1","SELE")

seu<-AddModuleScore_UCell(seu,features = signatures, name = "", ncores=8)
```

View signatures
```{r fig.height=12, fig.width=20}
FeaturePlot(seu, features=names(signatures)) & theme(aspect.ratio = 1,
  axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())
```

```{r}
VlnPlot(seu,features=names(signatures),
        group.by = "seurat_clusters", stack = T,flip = T,assay = "RNA")+ NoLegend()
```

```{r fig.height=12, fig.width=20}
FeaturePlot(seu, features=c("PTCH1","GLI1","GLI2","HHIP","MYCN","SPON2")) & theme(aspect.ratio = 1,
  axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

FeaturePlot(seu, features=c("RGS5","KCNE4","CPE","PTPRC")) & theme(aspect.ratio = 1,
  axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

FeaturePlot(seu, features=c("LYZ","SPI1","CSF1R","CD68","APOE","FCN1","S100A9")) & theme(aspect.ratio = 1,
  axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

FeaturePlot(seu, features=c("KRT1","KRT10","GATA3","KRTDAP","KRT5","KRT14")) & theme(aspect.ratio = 1,
  axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

```


```{r}
DimPlot(seu, reduction = "umap", group.by = "seurat_clusters", label=T) + theme(aspect.ratio = 1)
```




Clusters by majority vote
```{r}
tab <- table(seu$seurat_clusters, seu$scGate_broad, useNA = "ifany")
tab_norm <- apply(tab, 1, function(x) {x/sum(x)})
tab_norm

majority <- apply(tab_norm, 2, function(x){
  names(which.max(x))
})

seu$annotation <- unname(majority[seu$seurat_clusters])

DimPlot(seu, group.by = "annotation", label = T, repel = T, label.size = 2) +
    theme(aspect.ratio = 1)
```
Focus on T cells and Keratinocytes
```{r}
which.cells <- colnames(seu)[seu$annotation %in% c("Tcell","Epithelial") | is.na(seu$annotation)]
sub <- subset(seu, cells=which.cells)
```

## Dimensionality reduction and clustering

```{r fig.width=12}
sub <- NormalizeData(sub) |> FindVariableFeatures(nfeatures=2000) |>
  ScaleData()

sub <- sub |> RunPCA(npcs=30) |> RunUMAP(reduction = "pca", dims = 1:30)

a <- DimPlot(sub, reduction = "umap", group.by = "Sample") + theme(aspect.ratio = 1)
b <- DimPlot(sub, reduction = "umap", group.by = "annotation") + theme(aspect.ratio = 1)

a | b
```

```{r}
sub <- FindNeighbors(object = sub, dims=1:30)
sub <- FindClusters(object = sub, resolution = 1)
DimPlot(sub, reduction = "umap", group.by = "seurat_clusters") + theme(aspect.ratio = 1)
```

View signatures
```{r fig.height=12, fig.width=20}
FeaturePlot(sub, features=names(signatures)) & theme(aspect.ratio = 1,
  axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())
```


## Get mean expression by cluster
```{r}
BCC_means<-get_means_signature(sub,"BCC")
NormKerat_means<-get_means_signature(sub,"NormKerat")
TCells_means<-get_means_signature(sub,"TCells")
Myeloid_means <- get_means_signature(sub,"MyeloidCells")
```

Now add annotation for normal keratinocytes and cancer cells based on signatures

Any doublets?
```{r}
plot(x=BCC_means, y=NormKerat_means)
plot(x=BCC_means, y=TCells_means)
plot(x=NormKerat_means, y=TCells_means)
plot(x=NormKerat_means, y=Myeloid_means)
```

```{r}
tcell_pass <- names(TCells_means)[TCells_means>0.15]
bcc_pass <- names(BCC_means)[BCC_means>0.15]
ker_pass <- names(NormKerat_means)[NormKerat_means>0.20]
myeloid_pass <- names(Myeloid_means)[Myeloid_means>0.15]

doublets <- unique(c(intersect(tcell_pass, ker_pass),
                     intersect(tcell_pass, bcc_pass),
                     intersect(myeloid_pass, bcc_pass),
                     intersect(myeloid_pass, ker_pass)))

sub$annotation[sub$seurat_clusters %in% ker_pass] <- "Normal_Kerat"
sub$annotation[sub$seurat_clusters %in% bcc_pass] <- "Cancer_cells"
sub$annotation[sub$seurat_clusters %in% doublets] <- "Doublets"
```

```{r}
a <- DimPlot(sub, reduction = "umap", group.by = "Sample") + theme(aspect.ratio = 1)
a <- rasterize(a, layers='Point', dpi=300)

ann <- DimPlot(sub, group.by = "annotation", label = T, repel = T, label.size = 2) +
    theme(aspect.ratio = 1)
ann <- rasterize(ann, layers='Point', dpi=300)

(a | ann) & theme(axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank())

ggsave("plots/CG_kerat_anno_umap.pdf", width=12, heigh=5)
```

```{r}
table(sub$Sample, sub$annotation, useNA = "ifany")
```

```{r}
labels.sub <- sub$annotation

seu$annotation[names(labels.sub)] <- unname(labels.sub)

notna <- colnames(seu)[!is.na(seu$annotation)]
seu <- subset(seu, cells = notna)

table(seu$Sample, seu$annotation, useNA = "ifany")
```

# Rename patient by BCCX

```{r fig.width=12}
new.ids <-  paste0("BCC", seq(from=8, to=15))
names(new.ids) <- unique(seu$Sample)

seu$patient_bcc <- unname(new.ids[seu$Sample])

Idents(seu) <- "Sample"
```

Plot UMAPs (downsample for visuals only)
```{r fig.width=12, fig.height=5}
toplot <- seu
Idents(toplot) <- "annotation"
toplot <- subset(toplot, downsample = 20000)

a <- DimPlot(toplot, reduction = "umap", group.by = "Sample") + theme(aspect.ratio = 1)
a <- rasterize(a, layers='Point', dpi=300)

ann <- DimPlot(toplot, group.by = "annotation", label = T, repel = T, raster=F,
               label.size = 3, cols=palette) +
    theme(aspect.ratio = 1)
ann <- rasterize(ann, layers='Point', dpi=300)

(a | ann) & theme(axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank())

ggsave("plots/CG_all_anno_umap.pdf", width=12, heigh=5)
```

```{r}
seu$annotation <- factor(seu$annotation, levels=c("Cancer_cells","Normal_Kerat","Melanocyte","Endothelial",
                                                  "Fibroblast","Mast","MoMacDC","Bcell","PlasmaCell","Tcell"))
genes <- c("PTCH1","GLI1","GLI2","SPON2","KRT14","KRT5","PMEL","PECAM1","COL1A1","CPA3",
           "SPI1","CD79A","JCHAIN","CD2")

DotPlot(seu, features=genes, group.by = "annotation", cols="RdBu", scale=T, col.max=2) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
ggsave("plots/CG_dotplot.pdf", height=4, width=7)
```

```{r}
Idents(seu) <- "annotation"
VlnPlot(seu,features=names(signatures),
        group.by = "annotation", cols=palette, fill.by = "ident", stack = T) + NoLegend()
ggsave("plots/CG_signature_violin.pdf", width=7, height=4)
```


Evaluate sample mixing - is it lower among cancer cells?
```{r}
library(scIntegrationMetrics)

meta <- seu@meta.data[, c("annotation", "Sample")]

tab <- table(meta$annotation, useNA = "ifany")
pass <- names(tab)[tab>100] 

notNA.cells <- rownames(meta)[!is.na(meta[, "annotation"]) & meta[,"annotation"] %in% pass]

object <- subset(seu, cells = notNA.cells)
meta <- meta[notNA.cells, ]

emb <- Embeddings(object, reduction = "pca")

#iLISI
iLISI <- compute_lisi(X = emb, meta_data = object@meta.data, 
            label_colnames = "Sample")[[1]]

meta$iLISI <- iLISI

#sort by increasing average iLISI
agg <- aggregate(meta$iLISI, list(meta$annotation), FUN=mean) 
agg.v <- agg$x
names(agg.v) <- agg$Group.1
agg.sorted <- sort(agg.v, decreasing = F)

meta$annotation <- factor(meta$annotation, levels=names(agg.sorted))

ggplot(meta, aes(x=annotation, y=iLISI)) + 
  geom_boxplot(outlier.size=0.1) + theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust =1, hjust=1))

ggsave("plots/CG_iLISI_per_celltype.pdf", height=3.5, width=5)
```

On subset only
```{r}
library(scIntegrationMetrics)

meta <- sub@meta.data[, c("annotation", "Sample")]

tab <- table(meta$annotation, useNA = "ifany")
pass <- names(tab)[tab>100] 

notNA.cells <- rownames(meta)[!is.na(meta[, "annotation"]) & meta[,"annotation"] %in% pass]

object <- subset(sub, cells = notNA.cells)
meta <- meta[notNA.cells, ]

emb <- Embeddings(object, reduction = "pca")

#iLISI
iLISI <- compute_lisi(X = emb, meta_data = object@meta.data, 
            label_colnames = "Sample")[[1]]

meta$iLISI <- iLISI

#sort by increasing average iLISI
agg <- aggregate(meta$iLISI, list(meta$annotation), FUN=mean) 
agg.v <- agg$x
names(agg.v) <- agg$Group.1
agg.sorted <- sort(agg.v, decreasing = F)

meta$annotation <- factor(meta$annotation, levels=names(agg.sorted))

ggplot(meta, aes(x=annotation, y=iLISI)) + 
  geom_boxplot(outlier.size=0.1) + theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust =1, hjust=1))

ggsave("plots/CG_iLISI_per_celltype_sub.pdf", height=3.5, width=4)
```

# Save

```{r} 
file <- "cache/BCC_CG_all_annotated.rds"
saveRDS(seu, file)
```

Save cancer cells to file
```{r}
tumor <- subset(seu, subset=annotation == "Cancer_cells")

file_cancer <- "cache/Tumor_CG.rds"
saveRDS(tumor, file_cancer)
```
