---
title: "Annotated NMF meta-programs"
author: "Massimo Andreatta"
date: "`r format(Sys.Date(),'%e de %B, %Y')`"
output: 
  rmdformats::downcute:
    lightbox: true
    thumbnails: false
    self_contained: true
    gallery: true
    code_folding: show
  pkgdown:
    as_is: true
---

```{r, include=FALSE, fig.width=16, fig.height=12}
renv::restore()
library(Seurat)
library(ggplot2)
library(SignatuR)
library(UCell)
library(patchwork)
library(tidyr)
library(dplyr)
library(RColorBrewer)
library(viridis)
library(fgsea)
library(msigdbr)
library(GeneNMF)
library(openxlsx)
library(ggcorrplot)
library(corrr)
```



## Set the paths

Read in source data
```{r}
data.path <- "cache/Tumor_combination_LY_CG.rds"
seu <- readRDS(data.path)
```

Read in meta-programs (MPs) learned from the data using the script `2-NMF.Rmd`
```{r}
nmf.genes <- readRDS("cache/NMF_meta_res.rds")
```


# Intepretation of gene programs

Also include Nod-Inf signature for control
```{r}
morphologySignatures <- read.csv("_aux/spatialSignatures.csv")
morphologySignatures <- as.list(morphologySignatures)
morphologySignatures <- lapply(morphologySignatures, function(x) x[x!=""])
# a different universe (1800-gene panel) required for this GSEA
```

# Intepretation of filtered MPs based on Gavish et al pan-cancer gene meta programs

Compare MPs to Gavish MPs

```{r}
query_signature_list <- nmf.genes$metaprograms.genes

geneUniverse <- rownames(seu)[seu@assays$RNA@meta.features$vst.variance>0.01] 
#geneUniverse <- rownames(seu) 
reference_signature_list <- read.csv("_aux/Tirosh_malignantAndEpithelial_meta_programs_2023-07-13.csv")
reference_signature_list <- as.list(reference_signature_list)
reference_signature_list <- lapply(reference_signature_list, function(x) x[x!=""])

# unknown gene symbols?
refGenes <- unique(unlist(reference_signature_list))
unknownGenes <- refGenes[!refGenes %in% rownames(seu)]
unknownGenes
length(unknownGenes)/length(refGenes)*100 
# 6% gene symbols not mapped/detected

reference_signature_list <- lapply(reference_signature_list, function(x) x[!x %in% unknownGenes])

pval.thr <- 0.01

top_p <- lapply(query_signature_list, function(query_signature) {
  fgRes <- fgsea::fora(pathways = reference_signature_list,
                       genes = query_signature,
                       universe = geneUniverse)
  
  fgRes <- fgRes[fgRes$padj <= pval.thr,]
  
  
})

for (metaprogram in names(query_signature_list)) {
  
  top_p[[metaprogram]] <- cbind(metaprogram, top_p[[metaprogram]])
  print(head(top_p[[metaprogram]],20))
  
  top_p[[metaprogram]] <- mutate(top_p[[metaprogram]], qscore = -log(padj, base=10))
  top_p[[metaprogram]] <- mutate(top_p[[metaprogram]], geneRatio = round(overlap/size,3))
  top_p[[metaprogram]]$pathway <- factor(top_p[[metaprogram]]$pathway,levels=top_p[[metaprogram]]$pathway)
}
```

```{r}
enrich.list <- list()
for (i in names(top_p)){
  enrich.list[[i]] <- cbind(MP=i,top_p[[i]][,c("pathway","overlap","size")])
}

enrich.df <- do.call(rbind, enrich.list)
enrich.df <- as.data.frame(enrich.df)

#remove NAs
#enrich.df <- enrich.df[!is.na(enrich.df$pathway),]

enrich.df$Overlap_pct <- round(enrich.df$overlap/enrich.df$size*100)
enrich.df$Overlap_pct[is.na(enrich.df$Overlap_pct)] <- 0
#enrich.df$minLogPadj[enrich.df$minLogPadj > 10] <- 10
#enrich.df <- enrich.df[!is.na(enrich.df$padj),]
enrich.df
# Heatmap 
ggplot(enrich.df, aes(MP, pathway, fill= Overlap_pct)) + geom_tile() +
  scale_fill_gradient(low="white", high="blue", limits=c(0,80)) + theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

ggsave("plots/heatmap.Gavish_MP_enrichmentRefGeneOverlap.pdf", width=4.5, height=5)

#Horizontal?
ggplot(enrich.df, aes(pathway, MP, fill= Overlap_pct)) + geom_tile() +
  scale_fill_gradient(low="white", high="blue", limits=c(0,80)) + theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
ggsave("plots/heatmap.Gavish_MP_enrichmentRefGeneOverlap_flip.pdf", width=8, height=3.5)

```


Compare MPs to annotated signatures in msigDB 
```{r}
#geneCounts <- apply(seu@assays$RNA@data,1,sum)
#geneUniverse <- names(geneCounts[geneCounts>0])
geneUniverse <- rownames(seu) # very similar genes
```


```{r}
enrichmentSignatures <- c(nmf.genes$metaprograms.genes)
top_p <- lapply(enrichmentSignatures, function(metaprogram) {
  runGSEA(metaprogram, universe=geneUniverse, category = "C5", subcategory = "GO:BP")
#  runGSEA(metaprogram, universe=rownames(seu), category = "C2", subcategory = "CP:WIKIPATHWAYS")
})
```


```{r}
min.overlap <- 3
for (metaprogram in names(enrichmentSignatures)) {
  
  tab <- top_p[[metaprogram]]
  tab <- tab[tab$overlap >= min.overlap,]
  tab <- cbind(metaprogram, tab)
  
  tab <- mutate(tab, qscore = -log(padj, base=10))
  tab <- mutate(tab, geneRatio = round(overlap/size,3))
  tab$pathway <- factor(tab$pathway,levels=tab$pathway)
  #tab <- mutate(tab, path = substr(pathway,1,40))
  top_p[[metaprogram]] <- tab
  print(head(tab,15))
}

```

```{r}
enrichDotplot <- list()
cols <- brewer.pal(9, 'Blues')[3:9]

for (metaprogram in names(enrichmentSignatures)) {

  maxsize <- max(top_p[[metaprogram]]$overlap)

  enrichDotplot[[metaprogram]] <- ggplot(top_p[[metaprogram]][1:10,], aes(x = qscore, y = pathway,
                                                                          color = geneRatio, size = overlap)) + 
  geom_point(stat = 'identity') +
    scale_colour_gradientn(colours = cols) + #lims(size = c(1,maxsize)) +
#    scale_size(range=c(1,10), breaks = seq(1, maxsize, 3)) +
  xlab("-log(adjusted p-value)") + ylab("path") + ggtitle(metaprogram) + 
  theme_bw()
}

enrichDotplot

wp <- wrap_plots(enrichDotplot,ncol = 2)
ggsave("plots/enrichmentPlots.pdf",wp, width = 18, height = 15)
```


Annotate these signatures
```{r}
signature_lgt <- 200
sigs <- lapply(nmf.genes$metaprograms.genes, function(x) {
  lgt <- min(length(x), signature_lgt)
  head(x, lgt)
})
```


Label MPs

```{r}
names(sigs)[1] <- "MP1.CC"
names(sigs)[2] <- "MP2"
names(sigs)[3] <- "MP3.ED"
names(sigs)[4] <- "MP4"
names(sigs)[5] <- "MP5"
names(sigs)[6] <- "MP6"
names(sigs)[7] <- "MP7.WR"
sigs <- sigs[!is.na(names(sigs))]
```

```{r}
sigs
write.xlsx(sigs, "_aux/markers.raw.MP.xlsx")
```

Define cells and samples based on these signatures
```{r}
seu <- AddModuleScore_UCell(seu, features = sigs, assay="RNA", ncores=4, name = "")
```

# Determine strongest markers associated with each MP

Calculate to 20% and bottom 20% scores for each MP
```{r}
library(ProjecTILs)
markers.MP.list <- list()
set.seed(1234)
for (i in names(sigs)){
  i_class <- paste0(i,"_class")
  seu@meta.data[[i_class]] <- NA
  my_high_val <- quantile(seu@meta.data[[i]],p=c(0.80))
  seu@meta.data[[i_class]][seu@meta.data[[i]] > my_high_val] <- "high"
  
  my_low_val <- quantile(seu@meta.data[[i]],p=c(0.20))
  seu@meta.data[[i_class]][seu@meta.data[[i]] < my_low_val] <- "low"
  
  table(seu@meta.data[[i_class]],seu$Sample)
  
  Idents(seu) <- seu@meta.data[[i_class]]
  markers.MP.list[[i]] <- ProjecTILs::FindAllMarkers.bygroup(seu, split.by = "Sample",
                                                             min.diff.pct=0.1, min.freq=0.7, logfc.threshold = 2)
  head(markers.MP.list[[i]][["high"]],n=10)
}

markers.MP.list.highGenes <- lapply(markers.MP.list,function(x) {
  x[["high"]][["Gene"]]
})


markers.MP.list.highGenes.intersect <- list()
for(i in seq_along(sigs)){
  markers.MP.list.highGenes.intersect[[i]] <- intersect(markers.MP.list.highGenes[[i]],sigs[[i]])
}
names(markers.MP.list.highGenes.intersect) <- names(markers.MP.list.highGenes)
markers.MP.list.highGenes.intersect

write.xlsx(markers.MP.list.highGenes.intersect, "_aux/markers.MP.xlsx")
```


Quantify similarity between MP markers:
```{r}
library(scales)
overlap_coef <- function(vec1, vec2) {
  intersection <- length(intersect(vec1, vec2))
  #min_size <- min(c(length(vec1), length(vec2)))
  min_size <- length(vec1)
  return(intersection / min_size)
}


num_vectors <- length(markers.MP.list.highGenes.intersect)
similarity_matrix <- matrix(NA, nrow = num_vectors, ncol = num_vectors)
max_overlap_coef <- rep(0,length(markers.MP.list.highGenes.intersect))

# Compute similarity between all pairs of vectors
for (i in 1:num_vectors) {
  for (j in 1:num_vectors) {
    if (i != j) {
      similarity_matrix[i, j] <- overlap_coef(markers.MP.list.highGenes.intersect[[i]],
                                              markers.MP.list.highGenes.intersect[[j]])
      if (similarity_matrix[i, j] > max_overlap_coef[i]) { max_overlap_coef[i] <- similarity_matrix[i, j]}  
    } else {
      similarity_matrix[i, j] <- 1  
    }
  }
}
rownames(similarity_matrix) <- names(markers.MP.list.highGenes.intersect)
colnames(similarity_matrix) <- names(markers.MP.list.highGenes.intersect)

similarity_matrix <- 100* similarity_matrix

print(similarity_matrix)

palette <- gradient_n_pal(RColorBrewer::brewer.pal(9, "Blues"))(seq(0, 1, length.out = 30))

p <- pheatmap::pheatmap(similarity_matrix, col = palette, scale = "none",
                        angle_col = 45, treeheight_row = 10, treeheight_col = 10)
p 

ggsave("plots/heatmap_geneOverlapCoeff_MP_markers.pdf", plot = p, width=5, height = 4)
```




We observe that in terms of MP marker genes overlap, MP7 and MP3 are highly similar.

```{r}
flattened <- unlist(markers.MP.list.highGenes.intersect)
duplicated.genes <- flattened[duplicated(flattened)]
duplicated.genes

markers.MP.list.highGenes.intersect.specific <- lapply(markers.MP.list.highGenes.intersect,
                                                       function(x) x[!x %in% duplicated.genes])

write.xlsx(markers.MP.list.highGenes.intersect.specific, "_aux/markers.MP.specific.xlsx")
```

Pseudobulk the counts based on donor-condition-celltype
```{r}
pseudo_seu <- AggregateExpression(seu, assays = "RNA", return.seurat = T, group.by = c("MP7.WR_class", "Sample"))
mat <- LayerData(pseudo_seu)[markers.MP.list.highGenes.intersect.specific$MP7.WR,]
pheatmap::pheatmap(mat, col= heat.colors(8), scale = "row")
```

```{r, fig.height=5}
pseudo_seu <- AggregateExpression(seu, assays = "RNA", return.seurat = T, group.by = c("MP3.ED_class", "Sample"))

mat <- LayerData(pseudo_seu)[markers.MP.list.highGenes.intersect$MP3.ED,]
pheatmap::pheatmap(mat, col= heat.colors(8), scale = "row")

mat <- LayerData(pseudo_seu)[markers.MP.list.highGenes.intersect.specific$MP3.ED,]
pheatmap::pheatmap(mat, col= heat.colors(8), scale = "row")
```

# Intepretation of filtered MPs based on Gavish et al pan-cancer gene meta programs

Compare MPs to Gavish MPs

```{r}
geneUniverse <- rownames(seu)[seu@assays$RNA@meta.features$vst.variance>0.01] 
#geneUniverse <- rownames(seu) 
reference_signature_list <- read.csv("_aux/Tirosh_malignantAndEpithelial_meta_programs_2023-07-13.csv")
reference_signature_list <- as.list(reference_signature_list)
reference_signature_list <- lapply(reference_signature_list, function(x) x[x!=""])

# unknown gene symbols?
refGenes <- unique(unlist(reference_signature_list))
unknownGenes <- refGenes[!refGenes %in% rownames(seu)]
unknownGenes
length(unknownGenes)/length(refGenes)*100 
# 6% gene symbols not mapped/detected

reference_signature_list <- lapply(reference_signature_list, function(x) x[!x %in% unknownGenes])

query_signature_list <- markers.MP.list.highGenes.intersect

pval.thr <- 0.01

top_p <- lapply(query_signature_list, function(query_signature) {
  fgRes <- fgsea::fora(pathways = reference_signature_list,
                       genes = query_signature,
                       universe = geneUniverse)
  
  fgRes <- fgRes[fgRes$padj <= pval.thr,]
  
  
})

for (metaprogram in names(query_signature_list)) {
  
  top_p[[metaprogram]] <- cbind(metaprogram, top_p[[metaprogram]])
  print(head(top_p[[metaprogram]],20))
  
  top_p[[metaprogram]] <- mutate(top_p[[metaprogram]], qscore = -log(padj, base=10))
  top_p[[metaprogram]] <- mutate(top_p[[metaprogram]], geneRatio = round(overlap/size,3))
  top_p[[metaprogram]]$pathway <- factor(top_p[[metaprogram]]$pathway,levels=top_p[[metaprogram]]$pathway)
}


```


```{r}
enrich.list <- list()
for (i in names(top_p)){
  enrich.list[[i]] <- cbind(MP=i,top_p[[i]][,c("pathway","padj")])
}

enrich.df <- do.call(rbind, enrich.list)
enrich.df <- as.data.frame(enrich.df)

#remove NAs
enrich.df <- enrich.df[!is.na(enrich.df$pathway),]

enrich.df$minLogPadj <- -log(enrich.df$padj)
enrich.df$minLogPadj[is.na(enrich.df$padj)] <- 0
enrich.df$minLogPadj[enrich.df$minLogPadj > 10] <- 10
#enrich.df <- enrich.df[!is.na(enrich.df$padj),]
enrich.df
# Heatmap 
ggplot(enrich.df, aes(MP, pathway, fill= minLogPadj)) + geom_tile() +
  scale_fill_gradient(low="white", high="blue", limits=c(0,10)) + theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

ggsave("plots/heatmap.Gavish_MP_enrichment_deg.pdf", width=4.8, height=5)
```
```{r}
enrich.list <- list()
for (i in names(top_p)){
  enrich.list[[i]] <- cbind(MP=i,top_p[[i]][,c("pathway","overlap","size")])
}

enrich.df <- do.call(rbind, enrich.list)
enrich.df <- as.data.frame(enrich.df)

#remove NAs
#enrich.df <- enrich.df[!is.na(enrich.df$pathway),]

enrich.df$Overlap_pct <- round(enrich.df$overlap/enrich.df$size*100)
enrich.df$Overlap_pct[is.na(enrich.df$Overlap_pct)] <- 0
#enrich.df$minLogPadj[enrich.df$minLogPadj > 10] <- 10
#enrich.df <- enrich.df[!is.na(enrich.df$padj),]
enrich.df
# Heatmap 
ggplot(enrich.df, aes(MP, pathway, fill= Overlap_pct)) + geom_tile() +
  scale_fill_gradient(low="white", high="blue", limits=c(0,80)) + theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

ggsave("plots/heatmap.Gavish_MP_enrichmentRefGeneOverlap_deg.pdf", width=4.5, height=5)

#Horizontal?
ggplot(enrich.df, aes(pathway, MP, fill= Overlap_pct)) + geom_tile() +
  scale_fill_gradient(low="white", high="blue", limits=c(0,80)) + theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
ggsave("plots/heatmap.Gavish_MP_enrichmentRefGeneOverlap_flip_deg.pdf", width=8, height=3.5)

```


```{r}
write.xlsx(top_p, "_aux/Enrichment_Gavish_MP.xlsx")
```


Compare MPs to annotated signatures in msigDB 
```{r}
geneUniverse <- rownames(seu)

enrichmentSignatures <- markers.MP.list.highGenes.intersect

top_p <- lapply(enrichmentSignatures, function(metaprogram) {
   runGSEA(metaprogram, universe=geneUniverse, category = "C5", subcategory = "GO:BP")
#  runGSEA(metaprogram, universe=geneUniverse, category = "C5", subcategory = "GO:MF")
})
```


```{r}
min.overlap <- 3
for (metaprogram in names(enrichmentSignatures)) {
  
  tab <- top_p[[metaprogram]]
  tab <- tab[tab$overlap >= min.overlap,]
  tab <- cbind(metaprogram, tab)
  
  tab <- mutate(tab, qscore = -log(padj, base=10))
  tab <- mutate(tab, geneRatio = round(overlap/size,3))
  tab$pathway <- factor(tab$pathway,levels=tab$pathway)
  #tab <- mutate(tab, path = substr(pathway,1,40))
  top_p[[metaprogram]] <- tab
  print(head(tab,15))
}

```

```{r fig.width=10}
enrichDotplot <- list()
cols <- brewer.pal(9, 'Blues')[3:9]

for (metaprogram in names(enrichmentSignatures)) {

  maxsize <- max(top_p[[metaprogram]]$overlap)

  enrichDotplot[[metaprogram]] <- ggplot(top_p[[metaprogram]][1:10,], aes(x = qscore, y = pathway,
                                                                          color = geneRatio, size = overlap)) + 
  geom_point(stat = 'identity') +
    scale_colour_gradientn(colours = cols) + #lims(size = c(1,maxsize)) +
#    scale_size(range=c(1,10), breaks = seq(1, maxsize, 3)) +
  xlab("-log(adjusted p-value)") + ylab("path") + ggtitle(metaprogram) + 
  theme_bw()
}

enrichDotplot

wp <- wrap_plots(enrichDotplot,ncol = 2)
ggsave("plots/enrichmentPlots_deg.pdf",wp, width = 21, height = 15)
```



Label MPs
```{r}
sigs <- markers.MP.list.highGenes.intersect
names(sigs)[1] <- "MP1.CC"
names(sigs)[2] <- "MP2"
names(sigs)[3] <- "MP3.ED"
names(sigs)[4] <- "MP4"
names(sigs)[5] <- "MP5"
names(sigs)[6] <- "MP6.STR"
names(sigs)[7] <- "MP7.WR"
sigs <- sigs[!is.na(names(sigs))]
```

```{r}
sigs
write.xlsx(sigs, "_aux/markers.MP.xlsx")
```

Remove previous MP scores and MP_class factors from Seurat object metadata
```{r}
col2rem <- grep("MP", colnames(seu[[]]))
names(seu@meta.data)[col2rem]
seu@meta.data <- seu@meta.data[,-col2rem]
```

Define cells and samples based on these signatures
```{r}
seu <- AddModuleScore_UCell(seu, features = sigs, assay="RNA", ncores=4, name = "")
```

Matrix of cell programs by cells
```{r}
matrix <- as.matrix(seu@meta.data[,names(sigs)])

#keep centering and sd factors
MPscale <- list()
MPscale$center <- apply(matrix, 2, mean)
MPscale$sd <- apply(matrix, 2, sd)

matrix.scaled <- matrix
for (r in 1:nrow(matrix)) {
  matrix.scaled[r,] <- (matrix[r,] - MPscale$center) / MPscale$sd 
}
```



```{r}
corr_matrix <- cor(matrix)
colnames(matrix) <- gsub("\\.\\S+$", "", colnames(matrix))

ggcorrplot(corr_matrix, method="circle", type="upper", hc.order = F)

ggsave("plots/MP_scores_correlation.pdf", height=3, width=4)
```

```{r}
#save in Seurat object; also save scaling factors
seu@misc$MP_signatures <- sigs
seu@misc$MP_scale <- MPscale
seu@misc$MP_signatures_specific <- markers.MP.list.highGenes.intersect.specific
```


Add signature scores as new reduction in Seurat object, and calculate UMAP within Seurat
```{r}
colnames(matrix.scaled) <- paste0("NMF_",seq(1, ncol(matrix.scaled)))
#New dim reduction
seu@reductions[["NMFsignatures"]] <- new("DimReduc",
                                         cell.embeddings = matrix.scaled,
                                         assay.used = "RNA",
                                         key = "NMF_",
                                         global = FALSE)
```


Reduce to 2D
```{r fig.height=4, fig.width=20}
set.seed(123)
seu <- RunUMAP(seu, reduction="NMFsignatures", dims=1:length(seu@reductions[["NMFsignatures"]]),
               metric = "euclidean", reduction.name = "umap_MP")

FeaturePlot(seu, features = names(sigs)[1:8], reduction = "umap_MP", ncol=7,
            pt.size = 4, raster = T) &
  scale_color_viridis(option="B") &
   theme(aspect.ratio = 1, axis.text=element_blank(), axis.ticks=element_blank(), axis.title = element_blank())

ggsave("plots/MP_umap_by_program.pdf", height = 3, width = 18)
#ggsave("plots/MP_umap_by_program.png", height = 6, width = 10)
```


By patient
```{r}
pat_palette <- brewer.pal(11, name="Paired")
DimPlot(seu, group.by = "Sample", reduction = "umap_MP", cols=pat_palette, pt.size = 4, raster = T) +
  theme(aspect.ratio = 1)

ggsave("plots/MP_umap_by_patient.pdf", height = 4, width = 6)
```


```{r fig.height=4, fig.width=20}
VlnPlot(seu, features = names(sigs), group.by = "Sample", pt.size = 0, ncol=7, cols=pat_palette)

ggsave("plots/MP_violin_bt_patient.pdf", height = 3, width = 18)
```


Save object
```{r}
saveRDS(seu, file="_aux/BCC_seurat_NMFprograms.rds")
```


